---
title: "sawr08"
author: "Robert A. Stevens"
date: "December 31, 2016"
output: html_document
---

```{r, comment=NA}
library(lattice)
library(sp)
library(gstat)
library(geoR)
library(MASS)
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 8 Interpolation and Geostatistics

## 8.1 Introduction

## 8.2 Exploratory Data Analysis

```{r, comment=NA}
data(meuse)
coordinates(meuse) <- c("x", "y")
spplot(meuse, "zinc", do.log = T)
bubble(meuse, "zinc", do.log = T, key.space = "bottom")
```

```{r, comment=NA}
xyplot(log(zinc) ~ sqrt(dist), as.data.frame(meuse))
zn.lm <- lm(log(zinc) ~ sqrt(dist), meuse)
meuse$fitted.s <- predict(zn.lm, meuse) - mean(predict(zn.lm, meuse))
meuse$residuals <- residuals(zn.lm)
spplot(meuse, c("fitted.s", "residuals"))
```

**Figure 8.1**. Zinc as a function of distance to river (left), and fitted-residual maps (fitted.s: mean subtracted) for the linear regression model of log zinc and square-root transformed distance to the river

## 8.3 Non-Geostatistical Interpolation Methods

```{r, comment=NA}
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
meuse.grid <- as(meuse.grid, "SpatialPixelsDataFrame")
```

### 8.3.1 Inverse Distance Weighted Interpolation

```{r, comment=NA}
idw.out <- idw(zinc ~ 1, meuse, meuse.grid, idp = 2.5)
as.data.frame(idw.out)[1:5, ]
```

### 8.3.2 Linear Regression

```{r, comment=NA}
zn.lm <- lm(log(zinc) ~ sqrt(dist), meuse)
meuse.grid$pred <- predict(zn.lm, meuse.grid)
meuse.grid$se.fit <- predict(zn.lm, meuse.grid, se.fit = TRUE)$se.fit
```

```{r, comment=NA}
meuse.lm <- krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid)
```

```{r, comment=NA}
meuse.tr2 <- krige(log(zinc) ~ 1, meuse, meuse.grid, degree = 2)
```

```{r, comment=NA}
lm(log(zinc) ~ I(x^2) + I(y^2) + I(x*y) + x + y, meuse)
```

```{r, comment=NA}
lm(log(zinc) ~ poly(x, y, 2), meuse)
```

## 8.4 Estimating Spatial Correlation: The Variogram

### 8.4.1 Exploratory Variogram Analysis

```{r, comment=NA}
hscat(log(zinc) ~ 1, meuse, (0:9) * 100)
```

```{r, comment=NA}
variogram(log(zinc) ~ 1, meuse, cloud = TRUE)
```

**Figure 8.2**. Lagged scatter plot for the log-zinc data in the meuse data set 

```{r, comment=NA}
plot(variogram(log(zinc) ~ 1, meuse))
```

```{r, comment=NA}
sel <- plot(variogram(zinc ~ 1, meuse, cloud = TRUE), digitize = TRUE)
plot(sel, meuse)
```

**Figure 8.3**. Variogram cloud (top) and sample variogram (bottom) for log-zinc data; numbers next to symbols refer to the value Nh in (8.4)

**Figure 8.4**. Interactively selected point pairs on the variogram cloud (left) and map of selected point pairs (right)

**Figure 8.5**. Sample variogram (bold) compared to 100 variograms for randomly reallocated data (grey lines)

### 8.4.2 Cutoff, Lag Width, Direction Dependence

```{r, comment=NA}
plot(variogram(log(zinc) ~ 1, meuse))
```

```{r, comment=NA}
plot(variogram(log(zinc) ~ 1, meuse, alpha = c(0, 45, 90, 135)))
```

```{r, comment=NA}
plot(variogram(log(zinc) ~ 1, meuse, cutoff = 1000, width = 50))
```

```{r, comment=NA}
variogram(log(zinc) ~ 1, meuse, boundaries = c(0, 50, 100, seq(250, 1500, 250)))
```

### 8.4.3 Variogram Modeling

```{r, comment=NA}
show.vgms()
show.vgms(model = "Mat", kappa.range = c(0.1, 0.2, 0.5, 1, 2, 5, 10), max = 10)
```

```{r, comment=NA}
vgm(1, "Sph", 300)
vgm(1, "Sph", 300, 0.5)
v1 <- vgm(1, "Sph", 300, 0.5)
v2 <- vgm(0.8, "Sph", 800, add.to = v1) 
v2
vgm(0.5, "Nug", 0)
```

```{r, comment=NA}
vgm()
```

```{r, comment=NA}
v <- variogram(log(zinc) ~ 1, meuse)
plot(v)
```

**Figure 8.6**. Sample variogram (plus) and fitted model (dashed line) 

```{r, comment=NA}
fit.variogram(v, vgm(1, "Sph", 800, 1))
```

```{r, comment=NA}
fit.variogram(v, vgm(1, "Sph", 10, 1))
```

```{r, comment=NA}
v.fit <- fit.variogram(v, vgm(1, "Sph", 10, 1))
if(attr(v.fit, "singular")) stop("singular fit")
```

```{r, comment=NA}
attr(v.fit, "SSErr")
```

Table 8.1. Values for argument fit.method in function fit.variogram

fit.method Weight  
---------- ----------------  
1          N[j]  
2          N[j]/{γ(h[j])}^2  
6          1  
7          N[j]/h[j]^2  

```{r, comment=NA}
v.eye <- eyefit(variog(as.geodata(meuse["zinc"]), max.dist = 1500)) 
ve.fit <- as.vgm.variomodel(v.eye[[1]])
```

```{r, comment=NA}
fit.variogram(v, vgm(1, "Sph", 800, 0.06), fit.sills = c(FALSE, TRUE))
```

```{r, comment=NA}
fit.variogram.reml(log(zinc) ~ 1, meuse, model = vgm(0.6, "Sph", 800, 0.06))
```

### 8.4.4 Anisotropy

```{r, comment=NA}
v.dir <- variogram(log(zinc) ~ 1, meuse, alpha = (0:3)*45)
v.anis <- vgm(0.6, "Sph", 1600, 0.05, anis = c(45, 0.3))
plot(v.dir, v.anis)
```

**Figure 8.7**. Directional sample variogram (plus) and fitted model (dashed line) for four directions (0 is North, 90 is East)

```{r, comment=NA}
plot(variogram(log(zinc) ~ 1, meuse, map = TRUE, cutoff = 1000, width = 100))
```

### 8.4.5 Multivariable Variogram Modeling

```{r, comment=NA}
g <- gstat(NULL, "logCd", log(cadmium) ~ 1, meuse) 
g <- gstat(g, "logCu", log(copper) ~ 1, meuse)
g <- gstat(g, "logPb", log(lead) ~ 1, meuse)
g <- gstat(g, "logZn", log(zinc) ~ 1, meuse)
g
vm <- variogram(g)
vm.fit <- fit.lmc(vm, g, vgm(1, "Sph", 800, 1))
plot(vm, vm.fit)
```

**Figure 8.8**. Direct variograms (diagonal) and cross variograms (off-diagonal) along with fitted linear model of coregionalization (dashed line)

```{r, comment=NA}
cor(as.data.frame(meuse)[c("cadmium", "copper", "lead", "zinc")])
```

### 8.4.6 Residual Variogram Modeling

```{r, comment=NA}
variogram(log(zinc) ~ sqrt(dist), meuse)
```

```{r, comment=NA}
f <- log(zinc) ~ sqrt(dist)
vt <- variogram(f, meuse)
vt.fit <- fit.variogram(vt, vgm(1, "Exp", 300, 1)) 
vt.fit
g.wls <- gstat(NULL, "log-zinc", f, meuse, model = vt.fit, set = list(gls = 1))
(variogram(g.wls)$gamma - vt$gamma)/mean(vt$gamma)
```

## 8.5 Spatial Prediction

### 8.5.1 Universal, Ordinary, and Simple Kriging

```{r, comment=NA}
lz.sk <- krige(log(zinc) ~ 1, meuse, meuse.grid, v.fit, beta = 5.9)
lz.ok <- krige(log(zinc) ~ 1, meuse, meuse.grid, v.fit)
lz.uk <- krige(log(zinc) ~ sqrt(dist), meuse, meuse.grid, vt.fit)
```

Variograms?
No - Trend functions given?
  No - Inverse distance weighted interpolation 
  Yes - (local) trend surface prediction
Yes - Simulations?
  No - Trend coefficients given?
    Yes - Simple (co)kriging
    No - Trend has only intercept?
      Yes - Ordinary (co)kriging 
      No - Universal (co)kriging
  Yes - indicators?
    No - Sequential Gaussian (co)simulation
      Trend coefficients given?
        Yes - "simple" 
        No - "universal"
    Yes - Sequential Indicator (co)simulation

**Figure 8.9**. Decision tree for the gstat predict method

### 8.5.2 Multivariable Prediction: Cokriging

```{r, comment=NA}
cok.maps <- predict(vm.fit, meuse.grid)
names(cok.maps)
```

```{r, comment=NA}
spplot.vcov(cok.maps)
```

```{r, comment=NA}
vm2.fit <- vm.fit
vm2.fit$model[[3]]$range = c(0, 900) 
predict(vm2.fit, meuse.grid)
```

```{r, comment=NA}
vm2.fit$set <- list(nocheck = 1)
x <- predict(vm2.fit, meuse.grid)
names(as.data.frame(x))
any(as.data.frame(x)[c(2, 4, 6, 8)] < 0)
```

**Figure 8.10**. Cokriging variances (diagonal) and covariances (off-diagonal)

### 8.5.3 Collocated Cokriging

```{r, comment=NA}
g.cc <- gstat(NULL, "log.zinc", log(zinc) ~ 1, meuse, model = v.fit)
meuse.grid$distn <- meuse.grid$dist - mean(meuse.grid$dist) + mean(log(meuse$zinc))
vd.fit <- v.fit
vov <- var(meuse.grid$distn)/var(log(meuse$zinc))
vd.fit$psill <- v.fit$psill*vov
g.cc <- gstat(g.cc, "distn", distn ~ 1, meuse.grid, nmax = 1, model = vd.fit, 
              merge = c("log.zinc", "distn"))
vx.fit <- v.fit
vx.fit$psill <- sqrt(v.fit$psill * vd.fit$psill) * cor(meuse$dist, log(meuse$zinc))
g.cc <- gstat(g.cc, c("log.zinc", "distn"), model = vx.fit)
x <- predict(g.cc, meuse.grid)
```

**Figure 8.11**. Predictions for collocated cokriging, ordinary kriging, and universal kriging

### 8.5.4 Cokriging Contrasts

```{r, comment=NA}
demo(pcb)
```

### 8.5.5 Kriging in a Local Neighborhood

### 8.5.6 Change of Support: Block Kriging

```{r, comment=NA}
lz.ok <- krige(log(zinc) ~ 1, meuse, meuse.grid, v.fit, block = c(40, 40))
```

```{r, comment=NA}
xy <- expand.grid(x = seq(-20, 20, 4), y = seq(-20, 20, 4))
xy <- xy[(xy$x^2 + xy$y^2) <= 20^2, ]
lz.ok <- krige(log(zinc) ~ 1, meuse, meuse.grid, v.fit, block = xy)
```

```{r, comment=NA}
lz.pols <- krige(log(zinc) ~ 1, meuse, meuse.polygons, v.fit)
```

```{r, comment=NA}
spsample(polygon, n = 500, type = "regular", offset = c(0.5, 0.5))
```

### 8.5.7 Stratifying the Domain

```{r, comment=NA}
meuse$part.a <- idw(part.a ~ 1, meuse.grid, + meuse, nmax = 1)$var1.pred
```

```{r, comment=NA}
meuse$part.a <- meuse.grid$part.a[overlay(meuse.grid, meuse)]
```

```{r, comment=NA}
x1 <- krige(log(zinc) ~ 1, 
            meuse[meuse$part.a == 0, ], 
            meuse.grid[meuse.grid$part.a == 0, ], 
            model = vgm(0.548, "Sph", 900, 0.0654), 
            nmin = 20, nmax = 40, maxdist = 1000)
x2 <- krige(log(zinc) ~ 1, 
            meuse[meuse$part.a == 1, ], 
            meuse.grid[meuse.grid$part.a == 1, ], 
            model = vgm(0.716, "Sph", 900), 
            nmin = 20, nmax = 40, maxdist = 1000)
lz.stk <- rbind(as.data.frame(x1), as.data.frame(x2)) 
coordinates(lz.stk) <- c("x", "y")
lz.stk <- as(x, "SpatialPixelsDataFrame")
spplot(lz.stk["var1.pred"], main = "stratified kriging predictions")
```

### 8.5.8 Trend Functions and their Coefficients

```{r, comment=NA}
g.tr <- gstat(formula = log(zinc) ~ sqrt(dist), data = meuse, model = v.fit)
predict(g.tr, meuse[1, ])
predict(g.tr, meuse[1, ], BLUE = TRUE)
```

```{r, comment=NA}
predict(g, meuse[1, ], BLUE = TRUE, debug = 32)
```

```{r, comment=NA}
meuse$Int <- rep(1, 155)
g.tr <- gstat(formula = log(zinc) ~ -1 + Int + sqrt(dist), data = meuse, model = v.fit)
rn <- c("Intercept", "beta1")
df <- data.frame(Int = c(0, 1), dist = c(1, 0), row.names = rn) 
spdf <- SpatialPointsDataFrame(SpatialPoints(matrix(0, 2, 2)), df)
spdf
predict(g.tr, spdf, BLUE = TRUE)
```

### 8.5.9 Non-Linear Transforms of the Response Variable

```{r, comment=NA}
boxcox(zinc ~ sqrt(dist), data = as.data.frame(meuse))
```

```{r, comment=NA}
meuse$zinc.ns <- qqnorm(meuse$zinc, plot.it = FALSE)$x
```

```{r, comment=NA}
ind.f <- I(zinc < 500) ~ 1
ind.fit <- fit.variogram(variogram(ind.f, meuse), vgm(1, "Sph", 800, 1))
ind.kr <- krige(ind.f, meuse, meuse.grid, ind.fit)
summary(ind.kr$var1.pred)
```

### 8.5.10 Singular Matrix Errors

```{r, comment=NA}
meuse.dup <- rbind(as.data.frame(meuse)[1, ], as.data.frame(meuse)) 
coordinates(meuse.dup) = ~x + y
krige(log(zinc) ~ 1, meuse.dup, meuse[1, ], v.fit)
```

```{r, comment=NA}
zd <- zerodist(meuse.dup)
zd
krige(log(zinc) ~ 1, meuse.dup[-zd[, 1], ], meuse[1, ], v.fit)
```

```{r, comment=NA}
setL <- list(cn_max = 1e+10)
krige(log(zinc) ~ 1, meuse.dup, meuse[1, ], v.fit, set = setL)
```

## 8.6 Model Diagnostics

### 8.6.1 Cross Validation Residuals

```{r, comment=NA}
sel100 <- sample(1:155, 100)
m.model <- meuse[sel100, ]
m.valid <- meuse[-sel100, ]
v100.fit <- fit.variogram(variogram(log(zinc) ~ 1, m.model), vgm(1, "Sph", 800, 1))
m.valid.pr <- krige(log(zinc) ~ 1, m.model, m.valid, v100.fit)
resid.kr <- log(m.valid$zinc) - m.valid.pr$var1.pred 
summary(resid.kr)
resid.mean <- log(m.valid$zinc) - mean(log(m.valid$zinc)) 
R2 <- 1 - sum(resid.kr^2)/sum(resid.mean^2)
R2
```

```{r, comment=NA}
m.valid.pr$res <- resid.kr
bubble(m.valid.pr, "res")
```

```{r, comment=NA}
nfold <- 3
part <- sample(1:nfold, 155, replace = TRUE) 
sel <- (part != 1)
m.model <- meuse[sel, ]
m.valid <- meuse[-sel, ]
```

```{r, comment=NA}
v.fit <- vgm(0.59, "Sph", 874, 0.04)
cv155 <- krige.cv(log(zinc) ~ 1, meuse, v.fit, nfold = 5)
bubble(cv155, "residual", main = "log(zinc): 5-fold CV residuals")
```

### 8.6.2 Cross Validation z-Scores

```{r, comment=NA}
summary(cv155)
```

**Figure 8.12**. Cross validation residuals for fivefold cross validation; symbol size denotes residual magnitude, positive residuals indicate under-prediction

### 8.6.3 Multivariable Cross Validation

```{r, comment=NA}
g.cv <- gstat.cv(g, nmax = 40)
```

### 8.6.4 Limitations to Cross Validation

```{r, comment=NA}
v1.fit <- vgm(0.591, "Sph", 897, 0.0507)
v2.fit <- vgm(0.591, "Sph", 897, add.to = vgm(0.0507, "Sph", 40))
```

```{r, comment=NA}
set.seed(13331)
cv155.1 <- krige.cv(log(zinc) ~ 1, meuse, v1.fit, nfold = 5)
set.seed(13331)
cv155.2 <- krige.cv(log(zinc) ~ 1, meuse, v2.fit, nfold = 5)
summary(cv155.1$residual - cv155.2$residual)
```

```{r, comment=NA}
b1 <- krige(log(zinc) ~ 1, meuse, meuse.grid, v1.fit, block = c(40, 40))$var1.var
b2 <- krige(log(zinc) ~ 1, meuse, meuse.grid, v2.fit, block = c(40, 40))$var1.var
summary((b1 - b2)/b1)
```

## 8.7 Geostatistical Simulation

### 8.7.1 Sequential Simulation

```{r, comment=NA}
lzn.sim <- krige(log(zinc) ~ 1, meuse, meuse.grid, v.fit, + nsim = 6, nmax = 40)
spplot(lzn.sim)
```

**Figure 8.13**. Six realizations of conditional Gaussian simulation for log-zinc

### 8.7.2 Non-Linear Spatial Aggregation and Block Averages

```{r, comment=NA}
nsim <- 1000
cutoff <- 500
grd <- overlay(meuse.grid, area.sp)
sel.grid <- meuse.grid[!is.na(grd), ]
lzn.sim <- krige(log(zinc) ~ 1, meuse, sel.grid, v.fit, nsim = nsim, nmax = 40)
res <- apply(as.data.frame(lzn.sim)[1:nsim], 2, function(x) mean(x > log(cutoff)))
hist(res, main = paste("fraction above", cutoff), xlab = NULL, ylab = NULL)
```

**Figure 8.14**. A non-rectangular area for which a non-linear aggregation is required (left) and distribution of the areal fraction with zinc concentration above 500 ppm

```{r, comment=NA}
bkr <- krige(log(zinc) ~ 1, meuse, area.sp, v.fit)
1 - pnorm(log(cutoff), bkr$var1.pred, sqrt(bkr$var1.var))
```

### 8.7.3 Multivariable and Indicator Simulation

```{r, comment=NA}
cok.sims <- predict(vm.fit, meuse.grid, nsim = 1000)
```

```{r, comment=NA}
table(meuse$soil)
s1.fit <- fit.variogram(variogram(I(soil == 1) ~ 1, meuse), vgm(1, "Sph", 800, 1))
s1.sim <- krige(I(soil == 1) ~ 1, meuse, meuse.grid, s1.fit, nsim = 6, indicators = TRUE, nmax = 40)
spplot(s1.sim)
```

**Figure 8.15**. Six realizations of conditional indicator simulation for soil type 1

## 8.8 Model-Based Geostatistics and Bayesian Approaches

## 8.9 Monitoring Network Optimization

```{r, comment=NA}
m1 <- sapply(1:155, function(x) mean(krige(log(zinc) ~ 1, meuse[-x, ], meuse.grid, v.fit)$var1.var))
which(m1 == min(m1))
```

```{r, comment=NA}
plot(sort(m1))
```

```{r, comment=NA}
cutoff <- 1000
f <- function(x) {
  kr = krige(log(zinc) ~ 1, meuse[-x, ], meuse.grid, v.fit)
  mean(abs(pnorm((kr$var1.pred - log(cutoff))/sqrt(kr$var1.var)) - 0.5))
}
m2 <- sapply(1:155, f) > which(m2 == max(m2))
```

**Figure 8.16**. Candidate points for removal. (Left) For mean kriging variance, (right) for delineating the 1,000ppm contour. (Open circles) 10% most favorite points, (closed circles) 10% least favorite points

## 8.10 Other R Packages for Interpolation and Geostatistics

### 8.10.1 Non-Geostatistical Interpolation

### 8.10.2 spatial

### 8.10.3 RandomFields

```{r, comment=NA}
PrintModelList()
```

### 8.10.4 geoR and geoRglm

```{r, comment=NA}
plot(variog(as.geodata(meuse["zinc"]), max.dist = 1500))
```

### 8.10.5 fields
