---
title: "ASDAwR2-02"
author: "Robert A. Stevens"
date: "May 26, 2015"
output: html_document
---

```{r, comment=NA}
library(sp)
library(maps)
library(maptools)
library(rgdal)
library(gpclib)
library(rgeos)
library(raster)
```

*Applied Spatial Data Analysis with R* (2nd Ed.)

by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 2 Classes for Spatial Data in R

## 2.1 Introduction

## 2.2 Classes and Methods in R

```{r, comment=NA}
pi*10^2
```

```{r, comment=NA}
"*"(pi, "^"(10, 2))
```

```{r, comment=NA}
pi*(1:10)^2
```

```{r, comment=NA}
x <- pi*10^2 
x
print(x)
print(x, digits = 12)
```

```{r, comment=NA}
class(x)
typeof(x)
```

```{r, comment=NA}
class(cars)
typeof(cars)
names(cars)
summary(cars)
```

```{r, comment=NA}
str(cars)
```

```{r, comment=NA}
class(dist ~ speed)
```

```{r, comment=NA}
lm(dist ~ speed, data = cars)
```

```{r, comment=NA}
cars$qspeed <- cut(cars$speed, breaks = quantile(cars$speed), include.lowest = TRUE)
is.factor(cars$qspeed)
par(mfrow = c(1, 2))
plot(dist ~ speed, data = cars, main = "numerical: scatterplot")
plot(dist ~ qspeed, data = cars, main = "factor: boxplots")
```

**Figure 2.1**. Plot methods for a formula with numerical (left panel) and factor (right panel) right-hand side variables

```{r, comment=NA}
lm(dist ~ qspeed, data = cars)
```

## 2.3 Spatial Objects

```{r, comment=NA}
getClass("Spatial")
```

```{r, comment=NA}
getClass("CRS")
```

```{r, comment=NA}
m <- matrix(c(0, 0, 1, 1), ncol = 2, dimnames = list(NULL, c("min", "max")))
crs <- CRS(projargs = as.character(NA))
crs
S <- Spatial(bbox = m, proj4string = crs)
S
```

```{r, comment=NA}
# Spatial(matrix(c(350, 85, 370, 95), ncol = 2, 
#                dimnames = list(NULL, c("min", "max"))), 
#         proj4string = CRS("+longlat"))
bb <- matrix(c(350, 85, 370, 95), ncol = 2, dimnames = list(NULL, c("min", "max")))
Spatial(bb, proj4string = CRS("+proj=longlat"))
```

## 2.4 SpatialPoints

```{r, comment=NA}
setwd("~/GitHub/GIS")
CRAN_df <- read.table("CRAN051001a.txt", header = TRUE)
CRAN_mat <- cbind(CRAN_df$long, CRAN_df$lat)
row.names(CRAN_mat) <- 1:nrow(CRAN_mat)
str(CRAN_mat)
```

```{r, comment=NA}
getClass("SpatialPoints")
```

```{r, comment=NA}
llCRS <- CRS("+proj=longlat +ellps=WGS84")
CRAN_sp <- SpatialPoints(CRAN_mat, proj4string = llCRS) 
summary(CRAN_sp)
```

### 2.4.1 Methods

```{r, comment=NA}
bbox(CRAN_sp)
```

```{r, comment=NA}
proj4string(CRAN_sp)
proj4string(CRAN_sp) <- CRS(as.character(NA))
proj4string(CRAN_sp)
proj4string(CRAN_sp) <- llCRS
```

```{r, comment=NA}
brazil <- which(CRAN_df$loc == "Brazil")
brazil
coordinates(CRAN_sp)[brazil, ]
```

```{r, comment=NA}
summary(CRAN_sp[brazil, ])
```

```{r, comment=NA}
south_of_equator <- which(coordinates(CRAN_sp)[, 2] < 0)
summary(CRAN_sp[-south_of_equator, ])
```

### 2.4.2 Data Frames for Spatial Point Data

```{r, comment=NA}
str(row.names(CRAN_df))
```

```{r, comment=NA}
CRAN_spdf1 <- SpatialPointsDataFrame(CRAN_mat, CRAN_df, 
                                     proj4string = llCRS, match.ID = TRUE)
CRAN_spdf1[4, ]
str(CRAN_spdf1$loc)
str(CRAN_spdf1[["loc"]])
```

```{r, comment=NA}
s <- sample(nrow(CRAN_df))
CRAN_spdf2 <- SpatialPointsDataFrame(CRAN_mat, CRAN_df[s, ], 
                                     proj4string = llCRS, 
                                     match.ID = TRUE)
all.equal(CRAN_spdf2, CRAN_spdf1)
CRAN_spdf2[4, ]
```

```{r, comment=NA, error=TRUE}
CRAN_df1 <- CRAN_df
row.names(CRAN_df1) <- sample(c(outer(letters, letters, paste, sep = "")),
                              nrow(CRAN_df1))
try(CRAN_spdf3 <- SpatialPointsDataFrame(CRAN_mat, CRAN_df1, proj4string = llCRS, match.ID = TRUE))
```

- **SpatialPointsDataFrame**
    + **SpatialPoints**
    + coords.nrs 
    + data

- **Spatial**
    + bbox 
    + proj4string

- **SpatialPoints**
    + coords
    + **Spatial**
    
- **data.frame**

**Figure 2.2**. Spatial points classes and their slots; arrows show subclass extensions

```{r, comment=NA}
getClass("SpatialPointsDataFrame")
```

```{r, comment=NA}
names(CRAN_spdf1)
str(model.frame(lat ~ long, data = CRAN_spdf1), give.attr = FALSE)
```

```{r, comment=NA}
CRAN_spdf4 <- SpatialPointsDataFrame(CRAN_sp, CRAN_df)
all.equal(CRAN_spdf4, CRAN_spdf2)
```

```{r, comment=NA}
CRAN_df0 <- CRAN_df
coordinates(CRAN_df0) <- CRAN_mat
proj4string(CRAN_df0) <- llCRS
all.equal(CRAN_df0, CRAN_spdf2)
str(CRAN_df0, max.level = 2)
```

```{r, comment=NA}
CRAN_df1 <- CRAN_df
names(CRAN_df1)
coordinates(CRAN_df1) <- c("long", "lat") 
proj4string(CRAN_df1) <- llCRS
str(CRAN_df1, max.level = 2)
```

```{r, comment=NA}
setwd("~/GitHub/GIS")
turtle_df <- read.csv("seamap105_mod.csv")
summary(turtle_df)
```

```{r, comment=NA}
timestamp <- as.POSIXlt(strptime(as.character(turtle_df$obs_date), 
                                 "%m/%d/%Y %H:%M:%S"), "GMT")
turtle_df1 <- data.frame(turtle_df, timestamp = timestamp) 
turtle_df1$lon <- ifelse(turtle_df1$lon < 0, turtle_df1$lon + 360, turtle_df1$lon)
turtle_sp <- turtle_df1[order(turtle_df1$timestamp), ]
coordinates(turtle_sp) <- c("lon", "lat")
proj4string(turtle_sp) <- CRS("+proj=longlat +ellps=WGS84")
```

```{r, comment=NA}
gshhs.c.b <- system.file("share/gshhs_c.b", package = "maptools")
gpclibPermit()
pac <- Rgshhs(gshhs.c.b, level = 1, xlim = c(130, 250), ylim = c(15, 60), verbose = FALSE)
plot(pac$SP, axes = TRUE, col = "khaki2", xaxs = "i", yaxs = "i")
plot(turtle_sp, add = TRUE)
m_rle <- rle(months(turtle_sp$timestamp))
clen <- cumsum(m_rle$lengths[-length(m_rle$lengths)]) - 1
crds <- coordinates(turtle_sp)
text(crds[clen, ], labels = m_rle$values[-1], pos = 3, offset = 1.5, srt = 45)
```

**Figure 2.3**. Westward movements of a captive-raised adult loggerhead turtle (Caretta caretta) from 10 August 1996 to 12 August 1997 

## 2.5 SpatialLines

```{r, comment=NA}
getClass("Line")
getClass("Lines")
```

```{r, comment=NA}
getClass("SpatialLines")
```

```{r, comment=NA}
japan <- map("world", "japan", plot = FALSE)
p4s <- CRS("+proj=longlat +ellps=WGS84")
SLjapan <- map2SpatialLines(japan, proj4string = p4s) 
str(SLjapan, max.level = 2)
```

- **SpatialLines**
    + *lines*
    + **Spatial**

- **Lines**
    + *Lines*
    + ID

- **Line**
    + coords

- **SpatialPolygons**
    + *polygons*
    + plotOrder
    + **Spatial**

- **Polygons**
    + *Polygons*
    + plotOrder
    + labpt 
    + ID
    + area

- **Polygon**
    + labpt 
    + area 
    + hole
    + ringDir 
    + coords

- **Spatial**
    + bbox 
    + proj4string

**Figure 2.4**. SpatialLines and SpatialPolygons classes and slots; thin arrows show subclass extensions, thick arrows the inclusion of lists of objects

```{r, comment=NA}
Lines_len <- sapply(slot(SLjapan, "lines"), function(x) length(slot(x, "Lines")))
table(Lines_len)
```

```{r, comment=NA}
volcano_sl <- ContourLines2SLDF(contourLines(volcano))
t(slot(volcano_sl, "data"))
```

```{r, comment=NA}
llCRS <- CRS("+proj=longlat +ellps=WGS84")
auck_shore <- MapGen2SL("auckland_mapgen.dat", llCRS) 
summary(auck_shore)
```

```{r, comment=NA}
lns <- slot(auck_shore, "lines")
islands_auck <- sapply(lns, function(x) {
  crds <- slot(slot(x, "Lines")[[1]], "coords")
  identical(crds[1,], crds[nrow(crds),])
})
islands_sl <- auck_shore[islands_auck]
list_of_Lines <- slot(islands_sl, "lines")
islands_sp <- SpatialPolygons(lapply(list_of_Lines, function(x) {
    Polygons(list(Polygon(slot(slot(x, "Lines")[[1]], "coords"))),
      ID=slot(x, "ID"))
  }),
  proj4string=CRS("+proj=longlat +ellps=WGS84"))

plot(auck_shore)
legend("bottomleft", legend = "a)", bty = "n")
plot(auck_shore)
plot(islands_sp, add = TRUE, col = "grey")
legend("bottomleft", legend = "b)", bty = "n")
```

**Figure 2.5**. Two maps of shorelines around Auckland: (a) line representation, (b) line representation over-plotted with islands converted to polygons and shaded. Note that Waiheke Island, the large island to the east, is not closed, and so not found as an island

## 2.6 SpatialPolygons

```{r, comment=NA}
lns <- slot(auck_shore, "lines")
table(sapply(lns, function(x) length(slot(x, "Lines"))))
islands_auck <- sapply(lns, function(x) {
  crds <- slot(slot(x, "Lines")[[1]], "coords") 
  identical(crds[1, ], crds[nrow(crds), ])
})
table(islands_auck)
```

```{r, comment=NA}
getClass("Polygon")
```

```{r, comment=NA}
getClass("Polygons")
```

```{r, comment=NA}
getClass("SpatialPolygons")
```

```{r, comment=NA}
islands_sl <- auck_shore[islands_auck]
list_of_Lines <- slot(islands_sl, "lines")
islands_sp <- SpatialPolygons(
  lapply(list_of_Lines, 
         function(x) {
           Polygons(list(Polygon(slot(slot(x, "Lines")[[1]], "coords"))), 
                    ID = slot(x, "ID"))
         }), 
  proj4string = CRS("+proj=longlat +ellps=WGS84"))
summary(islands_sp)
slot(islands_sp, "plotOrder")
order(sapply(slot(islands_sp, "polygons"), function(x) slot(x, "area")), 
      decreasing = TRUE)
```

### 2.6.1 SpatialPolygonsDataFrame Objects

```{r, comment=NA}
state.map <- map("state", plot = FALSE, fill = TRUE)
IDs <- sapply(strsplit(state.map$names, ":"), function(x) x[1]) 
state.sp <- map2SpatialPolygons(state.map, IDs = IDs,
  proj4string = CRS("+proj=longlat +ellps=WGS84"))
```

```{r, comment=NA}
sat <- read.table("state.sat.data_mod.txt", row.names = 5, header = TRUE)
str(sat)
id <- match(row.names(sat), row.names(state.sp))
row.names(sat)[is.na(id)]
sat1 <- sat[!is.na(id), ]
state.spdf <- SpatialPolygonsDataFrame(state.sp, sat1)
str(slot(state.spdf, "data"))
str(state.spdf, max.level = 2)
```

```{r, comment=NA, error=FALSE}
rownames(sat1)[2] <- "Arizona"
try(SpatialPolygonsDataFrame(state.sp, sat1))
```

```{r, comment=NA}
DC <- "district of columbia"
not_dc <- !(row.names(slot(state.spdf, "data")) == DC) 
state.spdf1 <- state.spdf[not_dc, ]
length(slot(state.spdf1, "polygons"))
summary(state.spdf1)
```

### 2.6.2 Holes and Ring Direction

```{r, comment=NA}
load("high.RData")
manitoulin_sp <- high$SP
length(slot(manitoulin_sp, "polygons"))
sapply(slot(slot(manitoulin_sp, "polygons")[[1]], "Polygons"), function(x) slot(x, "hole"))
sapply(slot(slot(manitoulin_sp, "polygons")[[1]], "Polygons"), function(x) slot(x, "ringDir"))
```

```{r, comment=NA}
manitoulin_sp <- createSPComment(manitoulin_sp)
sapply(slot(manitoulin_sp, "polygons"), comment)

plot(manitoulin_sp, pbg = "lightsteelblue2", col = "khaki2", usePolypath = FALSE)
text(t(sapply(slot(slot(manitoulin_sp, "polygons")[[1]], "Polygons"), 
              function(x) slot(x, "labpt")))[-c(1,2),], 
     label = high$polydata$level[-c(1,2)], col = "black", font = 2)
cmt <- unlist(strsplit(sapply(slot(manitoulin_sp, "polygons"), comment), " "))
plot(manitoulin_sp, pbg = "lightsteelblue2", col = "khaki2", usePolypath = FALSE)
text(t(sapply(slot(slot(manitoulin_sp, "polygons")[[1]], "Polygons"), 
              function(x) slot(x, "labpt")))[-c(1, 2), ], 
     label = cmt[-c(1, 2)], col = "black", font = 2)
```

**Figure 2.6**. The northern, Canadian shore of Lake Huron, including Manitoulin Island and lakes on the island; islands (light grey) and lakes on islands (dark grey) are marked with their GSHHS levels

## 2.7 SpatialGrid and SpatialPixel Objects

```{r, comment=NA}
getClass("GridTopology")
```

```{r, comment=NA}
bb <- bbox(manitoulin_sp)
bb
cs <- c(0.01, 0.01)
cc <- bb[, 1] + (cs/2)
cd <- ceiling(diff(t(bb))/cs)
manitoulin_grd <- GridTopology(cellcentre.offset = cc, 
                               cellsize = cs, 
                               cells.dim = cd)
manitoulin_grd
```

```{r, comment=NA}
getClass("SpatialGrid")
```

```{r, comment=NA}
p4s <- CRS(proj4string(manitoulin_sp))
manitoulin_SG <- SpatialGrid(manitoulin_grd, proj4string = p4s) 
summary(manitoulin_SG)
```

```{r, comment=NA, error=TRUE}
#auck_el1 <- readGDAL("70042108.tif")
load("auck_el1.RData")
is.na(auck_el1$band1) <- auck_el1$band1 <= 0
load("auck_gshhs.RData")

class(auck_el1)
slot(auck_el1, "grid")
slot(auck_el1, "bbox")
object.size(auck_el1)
object.size(slot(auck_el1, "data"))
```

```{r, comment=NA}
r <- raster("70042108.tif")

class(r)
inMemory(r)
object.size(r)
cellStats(r, max)
cellStats(r, min)
inMemory(r)

out <- raster(r)
bs <- blockSize(out)
out <- writeStart(out, filename = tempfile(), overwrite = TRUE)
for (i in 1:bs$n) {
    v <- getValues(r, row = bs$row[i], nrows = bs$nrows[i])
    v[v <= 0] <- NA
    writeValues(out, v, bs$row[i])
}
out <- writeStop(out)
cellStats(out, min)
cellStats(out, max)
inMemory(out)

plot(out, col = terrain.colors(100))
plot(auck_gshhs, add = TRUE)
```

**Figure 2.7**. SRTM elevation data in meters for the Auckland isthmus over-plotted with an excerpt from the GSHHS full resolution shoreline, including two lakes – there are detailed differences stemming from the very different technologies underlying the two data sources. A transect is marked for later use

```{r, comment=NA}
r1 <- as(out, "SpatialGridDataFrame")
summary(r1)
r2 <- as(r1, "RasterLayer")
summary(r2)
```

```{r, comment=NA}
is.na(auck_el1$band1) <- auck_el1$band1 <= 0
summary(auck_el1$band1)
```

- **SpatialGridDataFrame**
    + **SpatialGrid**
    + **data**

- **SpatialGrid**
    + **grid**
    + grid.index
    + **SpatialPoints**

- **GridTopology**
    + cellcentre.offset 
    + cellsize 
    + cells.dim
    
- **SpatialPixelsDataFrame**
    + **SpatialPixels**
    + **data**

- **SpatialPixels**
    + **grid**
    + grid.index
    + **SpatialPoints**

- **Spatial**
    + bbox 
    + proj4string

- **SpatialPoints**
    + coords
    + **Spatial**

- **data.frame**

**Figure 2.8**. SpatialGrid and SpatialPixel classes and their slots; arrows show subclass extensions

```{r, comment=NA}
#auck_el2 <- as(auck_el1, "SpatialPixelsDataFrame") 
load("auck_el2.RData")

object.size(auck_el2)
object.size(slot(auck_el2, "grid.index"))
object.size(slot(auck_el2, "coords"))
sum(is.na(auck_el1$band1)) + nrow(slot(auck_el2, "coords"))
prod(slot(slot(auck_el2, "grid"), "cells.dim"))
```

```{r, comment=NA}
auck_el_500 <- auck_el2[auck_el2$band1 > 500, ]
summary(auck_el_500)
object.size(auck_el_500)
```

```{r, comment=NA}
data(meuse.grid)
mg_SP <- SpatialPoints(cbind(meuse.grid$x, meuse.grid$y)) 
summary(mg_SP)
mg_SPix0 <- SpatialPixels(mg_SP)
summary(mg_SPix0)
prod(slot(slot(mg_SPix0, "grid"), "cells.dim"))
```

```{r, comment=NA}
mg_SPix1 <- as(mg_SP, "SpatialPixels")
summary(mg_SPix1)
```
