---
title: "sawr06"
author: "Robert A. Stevens"
date: "November 17, 2016"
output: html_document
---

```{r, comment=NA}
library(sp)
library(lattice)
library(gstat)
library(rgdal)
library(xts)
library(spacetime)
library(maps)
library(maptools)
library(RColorBrewer)
library(plm)
setwd("~/GitHub/GIS")
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 6 Customizing Spatial Data Classes and Methods

## 6.1 Programming with Classes and Methods

```{r, comment=NA}
myfun <- function(x) {
  x + 2
}
```

```{r, comment=NA}
myfun(1:3)
```

```{r, comment=NA}
myfun(x = 1:3)
```

```{r, comment=NA}
plotXplus2Yminus3 <- function(x, y, ...) {
  plot(x = x + 2, y = y - 3, ...)
}
```

```{r, comment=NA}
methods("plot")
```

```{r, comment=NA}
showMethods("plot")
```

### 6.1.1 S3-Style Classes and Methods

```{r, comment=NA}
x <- rnorm(10)
class(x) <- "foo" 
x
```

```{r, comment=NA}
plot.foo <- function(x, y, ...) {
  plot.default(x, type = "l", ...)
}
```

```{r, comment=NA}
class(x) <- c("foo", "bar")
plot(x)
```

```{r, comment=NA}
data(meuse)
class(meuse)
class(lm(log(zinc) ~ sqrt(dist), meuse))
```

### 6.1.2 S4-Style Classes and Methods

```{r, comment=NA}
setClass("CRS", representation(projargs = "character")) 
setClass("Spatial", representation(bbox = "matrix",
  proj4string = "CRS"), validity <- function(object) { 
  bb <- bbox(object)
  if (!is.matrix(bb))
    return("bbox should be a matrix")
  n <- dimensions(object)
  if(n < 2)
    return("spatial.dimension should be 2 or more") 
  if(any(is.na(bb)))
    return("bbox should never contain NA values")
  if(any(!is.finite(bb)))
    return("bbox should never contain infinite values")
  if(any(bb[, "max"] < bb[, "min"]))
    return("invalid bbox: max < min")
  TRUE
})
```

```{r, comment=NA}
isGeneric("show")
```

```{r, comment=NA}
setGeneric("bbox", function(obj) standardGeneric("bbox"))
setMethod("bbox", signature = "Spatial", function(obj) obj@bbox)
```

## 6.2 Animal Track Data in Package Trip

```{r, comment=NA}
setClass("trip", representation("SpatialPointsDataFrame",
  TOR.columns = "character"), validity <- function(object) { 
  if(length(object@TOR.columns) != 2)
    stop("Time/id column names must be of length 2") 
  if(!all(object@TOR.columns %in% names(object@data)))
    stop("Time/id columns must be present in attribute table") 
  TRUE
})
showClass("trip")
```

### 6.2.1 Generic and Constructor Functions

```{r, comment=NA}
trip.default <- function(obj, TORnames) { 
  if(!is(obj, "SpatialPointsDataFrame"))
    stop("trip only supports SpatialPointsDataFrame")
  if (is.numeric(TORnames))
    TORnames <- names(obj)[TORnames]
  new("trip", obj, TOR.columns = TORnames)
}
if (!isGeneric("trip")) setGeneric("trip", function(obj, TORnames) standardGeneric("trip"))
setMethod("trip", signature(obj = "SpatialPointsDataFrame", TORnames = "ANY"), trip.default)
```

```{r, comment=NA}
turtle <- read.csv("seamap105_mod.csv")
timestamp <- as.POSIXlt(strptime(as.character(turtle$obs_date), "%m/%d/%Y %H:%M:%S"), "GMT")
turtle <- data.frame(turtle, timestamp = timestamp) 
turtle$lon <- ifelse(turtle$lon < 0, turtle$lon + 360, turtle$lon)
turtle <- turtle[order(turtle$timestamp), ] 
coordinates(turtle) <- c("lon", "lat")
proj4string(turtle) <- CRS("+proj=longlat +ellps=WGS84") 
turtle$id <- c(rep(1, 200), rep(2, nrow(coordinates(turtle)) - 200))
turtle_trip <- trip(turtle, c("timestamp", "id"))
summary(turtle_trip)
```

### 6.2.2 Methods for Trip Objects

```{r, comment=NA}
summary.trip <- function(object, ...) {
  cat("Object of class \"trip\"\nTime column: ")
  print(object@TOR.columns[1])
  cat("Identifier column: ") 
  print(object@TOR.columns[2]) 
  print(summary(as(object, "Spatial"))) 
  print(summary(object@data))
}
setMethod("summary", "trip", summary.trip)
summary(turtle_trip)
```

```{r, comment=NA}
setGeneric("lines", function(x, ...) standardGeneric("lines"))
setMethod("lines", signature(x = "trip"), function(x, ..., col = NULL) {
  tor <- x@TOR.columns
  if(is.null(col)) {
    l <- length(unique(x[[tor[2]]]))
    col <- hsv(seq(0, 0.5, length = l))
  }
  coords <- coordinates(x)
  lx <- split(1:nrow(coords), x[[tor[2]]])
  for(i in 1:length(lx)) lines(coords[lx[[i]], ],
    col = col[i], ...)
})
```

## 6.3 Multi-Point Data: SpatialMultiPoints

```{r, comment=NA}
setClass("SpatialMultiPoints", representation("SpatialLines"),
  validity <- function(object) {
    if(any(unlist(lapply(object@lines,
      function(x) length(x@Lines))) != 1))
      stop("Only Lines objects with one Line element")
    TRUE
})
SpatialMultiPoints <- function(object) new("SpatialMultiPoints", object)
```

```{r, comment=NA}
n <- 5
set.seed(1)
x1 <- cbind(rnorm(n), rnorm(n, 0, 0.25)) 
x2 <- cbind(rnorm(n), rnorm(n, 0, 0.25)) 
x3 <- cbind(rnorm(n), rnorm(n, 0, 0.25)) 
L1 <- Lines(list(Line(x1)), ID = "mp1") 
L2 <- Lines(list(Line(x2)), ID = "mp2") 
L3 <- Lines(list(Line(x3)), ID = "mp3") 
s <- SpatialLines(list(L1, L2, L3))
smp <- SpatialMultiPoints(s)
```

```{r, comment=NA}
plot.SpatialMultiPoints <- function(x, ..., pch = 1:length(x@lines), col = 1, cex = 1) {
  n <- length(x@lines)
  if(length(pch) < n)
    pch <- rep(pch, length.out = n)
  if(length(col) < n)
    col <- rep(col, length.out = n)
  if(length(cex) < n)
    cex <- rep(cex, length.out = n)
  plot(as(x, "Spatial"), ...)
  for(i in 1:n) points(x@lines[[i]]@Lines[[1]]@coords,
    pch = pch[i], col = col[i], cex = cex[i])
}
setMethod("plot", signature(x = "SpatialMultiPoints", y = "missing"), 
          function(x, y, ...) plot.SpatialMultiPoints(x, ...))
```

```{r, comment=NA}
cName <- "SpatialMultiPointsDataFrame"
setClass(cName, representation("SpatialLinesDataFrame"), 
  validity <- function(object) {
    lst <- lapply(object@lines, function(x) length(x@Lines)) 
    if(any(unlist(lst) != 1))
      stop("Only Lines objects with single Line") 
    TRUE
})
SpatialMultiPointsDataFrame <- function(object) {
  new("SpatialMultiPointsDataFrame", object)
}
```

```{r, comment=NA}
df <- data.frame(x1 = 1:3, x2 = c(1, 4, 2), row.names = c("mp1", "mp2", "mp3"))
smp_df <- SpatialMultiPointsDataFrame(SpatialLinesDataFrame(smp, df))
setMethod("plot", signature(x = "SpatialMultiPointsDataFrame", y = "missing"), 
          function(x, y, ...) plot.SpatialMultiPoints(x, ...))
grys <- c("grey10", "grey40", "grey80")
plot(smp_df, col = grys[smp_df[["x1"]]], pch = smp_df[["x2"]], cex = 2, axes = TRUE)
```

**Figure 6.1**. Plot of the SpatialMultiPointsDataFrame object

## 6.4 Hexagonal Grids

```{r, comment=NA}
data(meuse.grid)
gridded(meuse.grid) = ~x + y
xx <- spsample(meuse.grid, type = "hexagonal", cellsize = 200)
class(xx)
```

```{r, comment=NA}
HexPts <- spsample(meuse.grid, type = "hexagonal", cellsize = 200)
spplot(meuse.grid["dist"], sp.layout = list("sp.points", HexPts, col = 1))
HexPols <- HexPoints2SpatialPolygons(HexPts)
df <- as.data.frame(meuse.grid)[overlay(meuse.grid, HexPts), ]
HexPolsDf <- SpatialPolygonsDataFrame(HexPols, df, match.ID = FALSE)
spplot(HexPolsDf["dist"])
```

**Figure 6.2**. Hexagonal points (left) and polygons (right)

```{r, comment=NA}
setClass("SpatialHexGrid", representation("SpatialPoints", dx = "numeric"), 
  validity <- function(object) {
    if(object@dx <= 0)
      stop("dx should be positive")
    TRUE
})
setClass("SpatialHexGridDataFrame", representation("SpatialPointsDataFrame", dx = "numeric"), 
  validity <- function(object) { 
    if(object@dx <= 0)
      stop("dx should be positive")
    TRUE
})
```

```{r, comment=NA}
HexPts <- spsample(meuse.grid, type = "hexagonal", cellsize = 200) 
Hex <- new("SpatialHexGrid", HexPts, dx = 200)
df <- as.data.frame(meuse.grid)[overlay(meuse.grid, Hex), ]
spdf <- SpatialPointsDataFrame(HexPts, df)
HexDf <- new("SpatialHexGridDataFrame", spdf, dx = 200)
```

```{r, comment=NA}
is(HexDf, "SpatialHexGrid")
setIs("SpatialHexGridDataFrame", "SpatialHexGrid")
is(HexDf, "SpatialHexGrid")
```

```{r, comment=NA}
setAs("SpatialHexGrid", "SpatialPolygons", function(from) HexPoints2SpatialPolygons(from, from@dx))
setAs("SpatialHexGridDataFrame", "SpatialPolygonsDataFrame", 
      function(from) SpatialPolygonsDataFrame(as(obj, "SpatialPolygons"), obj@data, match.ID = FALSE))
```

```{r, comment=NA}
setMethod("plot", signature(x = "SpatialHexGrid", y = "missing"), 
          function(x, y, ...) plot(as(x, "SpatialPolygons"), \vspace*{-3pt} ...))
setMethod("spplot", signature(obj = "SpatialHexGridDataFrame"), 
          function(obj, ...) 
            spplot(SpatialPolygonsDataFrame(as(obj, "SpatialPolygons"), 
                                            obj@data, match.ID = FALSE), ...))
setMethod("spsample", "SpatialHexGrid", 
          function(x, n, type, ...) spsample(as(x, "SpatialPolygons"), n = n, type = type, ...))
setMethod("overlay", c("SpatialHexGrid", "SpatialPoints"), 
          function(x, y, ...) overlay(as(x, "SpatialPolygons"), y))
```

```{r, comment=NA}
spplot(meuse.grid["dist"], sp.layout = list("sp.points", + Hex, col = 1))
spplot(HexDf["dist"])
```

```{r, comment=NA}
as(HexDf, "data.frame")
```

```{r, comment=NA}
bbox(Hex)
bbox(as(Hex, "SpatialPolygons"))
```

## 6.5 Spatio-Temporal Grids

```{r, comment=NA}
n <- 10
x <- data.frame(expand.grid(x1 = 1:n, x2 = 1:n, x3 = 1:n), z = rnorm(n^3))
coordinates(x) <- ~x1 + x2 + x3
gridded(x) <- TRUE
fullgrid(x) <- TRUE
summary(x)
```

```{r, comment=NA}
setClass("SpatialTimeGrid", "SpatialGrid", validity <- function(object) {
  stopifnot(dimensions(object) == 3)
  TRUE
})
```

```{r, comment=NA}
setClass("SpatialTimeGridDataFrame", "SpatialGridDataFrame", validity <- function(object) {
  stopifnot(dimensions(object) == 3)
  TRUE
})
setIs("SpatialTimeGridDataFrame", "SpatialTimeGrid") 
x <- new("SpatialTimeGridDataFrame", x)
```

```{r, comment=NA}
summary.SpatialTimeGridDataFrame <- function(object, ...) {
  cat("Object of class SpatialTimeGridDataFrame\n") 
  x <- gridparameters(object)
  t0 <- ISOdate(1970, 1, 1, 0, 0, 0) 
  t1 <- t0 + x[3, 1]
  cat(paste("first time step:", t1, "\n")) 
  t2 <- t0 + x[3, 1] + (x[3, 3] - 1)*x[3, 2] 
  cat(paste("last time step: ", t2, "\n")) 
  cat(paste("time step: ", x[3, 2], "\n")) 
  summary(as(object, "SpatialGridDataFrame"))
}
setMethod("summary", "SpatialTimeGridDataFrame", summary.SpatialTimeGridDataFrame)
summary(x)
```

```{r, comment=NA}
subs.SpatialTimeGridDataFrame <- function(x, i, j, ..., drop = FALSE) {
  t <- coordinates(x)[, 3] + ISOdate(1970, 1, 1, 0, 0, 0)
  if(missing(j))
    j <- TRUE
  sel <- t %in% i
  if(!any(sel))
    stop("selection results in empty set")
  fullgrid(x) <- FALSE
  if(length(i) > 1) {
    x <- x[i = sel, j = j, ...]
    fullgrid(x) <- TRUE
    as(x, "SpatialTimeGridDataFrame")
  }
  else {
    gridded(x) <- FALSE
    x <- x[i = sel, j = j, ...]
    cc <- coordinates(x)[, 1:2]
    p4s <- CRS(proj4string(x))
    SpatialPixelsDataFrame(cc, x@data, proj4string = p4s)
  }
}
setMethod("[", c("SpatialTimeGridDataFrame", "POSIXct", "ANY"), subs.SpatialTimeGridDataFrame)
t1 <- as.POSIXct("1970-01-01 0:00:03", tz = "GMT") 
t2 <- as.POSIXct("1970-01-01 0:00:05", tz = "GMT") 
summary(x[c(t1, t2)])
summary(x[t1])
```

```{r, comment=NA}
spplot.stgdf <- function(obj, zcol = 1, ..., format = NULL) {
  if (length(zcol) != 1)
    stop("can only plot a single attribute")
  if (is.null(format))
    format <- "%Y-%m-%d %H:%M:%S"
  cc <- coordinates(obj)
  df <- unstack(data.frame(obj[[zcol]], cc[, 3]))
  ns <- as.character(coordinatevalues(getGridTopology(obj))[[3]] + ISOdate(1970, 1, 1, 0, 0, 0), 
                     format = format) 
  cc2d <- cc[cc[, 3] == min(cc[, 3]), 1:2]  
  obj <- SpatialPixelsDataFrame(cc2d, df) 
  spplot(obj, names.attr = ns, ...)
}
setMethod("spplot", "SpatialTimeGridDataFrame", spplot.stgdf)
```

```{r, comment=NA}
trellis.par.set(canonical.theme(color = FALSE))
spplot(x, format = "%H:%M:%S", as.table = TRUE, cuts = 6, 
       col.regions = grey.colors(7, 0.55, 0.95, 2.2))
```

```{r, comment=NA}
`?`(as.character.POSIXt)
```

## 6.6 Analyzing Spatial Monte Carlo Simulations

**Figure 6.3**. spplot for an object of class SpatialTimeGridDataFrame, filled with random numbers

```{r, comment=NA}
data(meuse)
coordinates(meuse) <- ~x + y
v <- vgm(0.5, "Sph", 800, 0.05)
sim <- krige(log(zinc) ~ 1, meuse, meuse.grid, v, nsim = 100, nmax = 30)
sim@data <- exp(sim@data)
```

```{r, comment=NA}
quantile.Spatial <- function(x, ..., byLayer = FALSE) {
  stopifnot("data" %in% slotNames(x))
  apply(x@data, ifelse(byLayer, 2, 1), quantile, ...)
}
```

```{r, comment=NA}
sim$lower <- quantile.Spatial(sim[1:100], probs = 0.025) 
sim$upper <- quantile.Spatial(sim[1:100], probs = 0.975)
```

```{r, comment=NA}
medians <- quantile.Spatial(sim[1:100], probs = 0.5, byLayer = TRUE)
hist(medians)
```

```{r, comment=NA}
fractionBelow <- function(x, q, byLayer = FALSE) { 
  stopifnot(is(x, "Spatial") || !("data" %in% slotNames(x)))
  apply(x@data < q, ifelse(byLayer, 2, 1), function(r) sum(r)/length(r))
}
over500 <- 1 - fractionBelow(sim[1:100], 200, byLayer = TRUE) 
summary(over500)
quantile(over500, c(0.025, 0.975))
```

## 6.7 Processing Massive Grids

```{r, comment=NA}
x <- readGDAL("70042108.tif", output.dim = c(120, 132)) 
x$band1[x$band1 <= 0] <- NA
spplot(x, col.regions = bpy.colors())
```

```{r, comment=NA}
x <- GDAL.open("70042108.tif")
class(x)
x.subs <- x[1:100, 1:100, 1]
class(x.subs)
gridparameters(x.subs)
```

```{r, comment=NA}
setClass("SpatialGDAL", 
         representation("Spatial", grid = "GridTopology", 
                        grod = "GDALReadOnlyDataset", name = "character"))
setClass("SpatialGDALWrite", "SpatialGDAL")
```

```{r, comment=NA}
x <- open.SpatialGDAL("70042108.tif")
nrows <- GDALinfo("70042108.tif")["rows"]
ncols <- GDALinfo("70042108.tif")["columns"]
xout <- copy.SpatialGDAL(x, "70042108out.tif")
bls <- 20
for (i in 1:(nrows/bls - 1)) {
  r <- 1 + (i - 1)*bls
  for(j in 1:(ncols/bls -1 )){
    c <- 1 + (j - 1)*bls
    x.in <- x[r:(r + bls), c:(c + bls)]
    xout[r:(r + bls), c:(c + bls)] <- x.in$band1 + 10
  }
  cat(paste("row-block", i, "\n"))
}
close(x)
close(xout)
```

```{r, comment=NA}
setMethod("[", "SpatialGDAL", function(x, i, j, ..., drop = FALSE) x@grod[i = i, j = j, ...])
setReplaceMethod("[", "SpatialGDALWrite", function(x, i, j, ..., value) {
  ...
})
```
