---
title: "ASDAwR2-01"
author: "Robert A. Stevens"
date: "July 10, 2017"
output: html_document
---

```{r, comment=NA}
library(maptools)
library(maps)
library(rgdal)
library(sp)
grey_gamma <- 2.2
```

*Applied Spatial Data Analysis with R* (2nd Ed.)

by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 1 Hello World: Introducing Spatial Data

## 1.1 Applied Spatial Data Analysis

## 1.2 Why Do We Use R

### 1.2.1 ... In General?

### 1.2.2 ... for Spatial Data Analysis?

**Figure 1.1** shows how the reception of sp classes has already influenced the landscape of contributed packages; interfacing other packages for handling and analyzing spatial data is usually simple as we see in Part II. The shaded nodes of the dependency graph are packages (co)-written and/or maintained by the authors of this book, and will be used extensively in the following chapters.

maptools 
rgdal
splancs
geoR
gstat 
spsurvey 
trip
aspace 
spdep 
spgwr 
surveillance 
GeoXp
spgrass6
GEOmap
ecespa
StatDA 
geoRglm 
simba
DCluster 
svcR 
BARD
RTOMO
VIM
sp

**Figure 1.1**. Tree of R contributed packages on CRAN depending on or importing sp directly or indirectly; others suggest sp or use it without declaration in their package descriptions (status as of 2008-04-06)

## 1.3 R and GIS 

### 1.3.1 What is GIS?

### 1.3.2 Service-Oriented Architectures

### 1.3.3 Further Reading on GIS

## 1.4 Types of Spatial Data

```{r, comment=NA}
setwd("~/GitHub/GIS")
volc.tab <- read.table("data1964al.xy")
volc <- SpatialPoints(volc.tab[c(2, 1)])
llCRS <- CRS("+proj=longlat +ellps=WGS84")
proj4string(volc) <- llCRS
prj_new <- CRS("+proj=moll +ellps=WGS84")
volc_proj <- spTransform(volc, prj_new)
wrld <- map("world", interior = FALSE, xlim = c(-179, 179), ylim = c(-89, 89), plot = FALSE)
wrld_p <- pruneMap(wrld, xlim = c(-179, 179))
wrld_sp <- map2SpatialLines(wrld_p, proj4string = llCRS)
wrld_proj <- spTransform(wrld_sp, prj_new)
#save(c("wrld_proj", "wrld_sp"), file = "hsd_data/wrld.RData")
#load("hsd_data/wrld.RData")
wrld_grd <- gridlines(wrld_sp, easts = c(-179,seq(-150, 150, 50), 179.5), 
                      norths = seq(-75, 75, 15), ndiscr = 100)
wrld_grd_proj <- spTransform(wrld_grd, prj_new)
at_sp <- gridat(wrld_sp, easts = 0, norths = seq(-75, 75, 15), offset = 0.3)
at_proj <- spTransform(at_sp, prj_new)
plot(wrld_proj, col = "grey50")
plot(wrld_grd_proj, add = TRUE, lty = 3, col = "grey50")
points(volc_proj, cex = 0.8, pch = 3, col = "blue")
text(coordinates(at_proj), pos=at_proj$pos, offset=at_proj$offset, 
     labels = parse(text = as.character(at_proj$labels)), cex = 0.6)
```

**Figure 1.2**. Volcanoes of the world, with last known eruption 1964 or later (+); source: National Geophysical Data Center

```{r, comment=NA}
data(volcano)
grys <- grey.colors(8, 0.55, 0.95, grey_gamma)
#layout(matrix(c(1, 2, 1, 3, 1, 4), 3, 2, byrow = TRUE), c(3, 1))
# a
image(volcano, axes = FALSE, col = grys, asp = 1, main = "a")
contour(volcano, add = TRUE)
box()
# b
image(volcano, axes = FALSE, col = 'white', asp = 1, main = "b")
x2 <- ContourLines2SLDF(contourLines(volcano))
plot(x2, add = TRUE)
box()
# c
image(volcano, axes = FALSE, col = 'white', asp = 1, main = "c")
plot(x2[x2$level == 140, ], add = TRUE)
box()
# d
image(volcano, axes = FALSE, col = grys, asp = 1, main = "d")
x3l1 <- coordinates(x2[x2$level == 160,])[[1]][[1]]
x3l2 <- coordinates(x2[x2$level == 160,])[[1]][[2]]
x3 <- SpatialPolygons(list(Polygons(list(Polygon(x3l1, hole = FALSE), 
                                        Polygon(x3l2, hole = TRUE)), ID = c("x"))))
plot(x3, col = '#FF8800', add = TRUE)
box()
```

**Figure. 1.3**. Maunga Whau (Mt Eden) is one of about 50 volcanoes in the Auckland volcanic field. (a) Topographic information (altitude, m) for Maunga Whau on a 10 × 10 m2 grid, (b) contour lines, (c) 140 m contour line: a closed polygon, (d) area above 160m (hashed): a polygon with a hole

## 1.5 Storage and Display

## 1.6 Applied Spatial Data Analysis

## 1.7 R Spatial Resources

### 1.7.1 Online Resources

### 1.7.2 Layout of the Book

From:  http://www.asdar-book.org/data2ed.php?chapter=0

# Chapter 1: Code and download bundle

**R Code**

```{r, comment=NA, eval=FALSE}
source("hello_mod.R", echo = TRUE)
```

```{r, comment=NA, eval=FALSE}
ASDAR_BOOK <- "http://www.asdar-book.org/book2ed"
chapters <- c("hello", "cm", "vis", "die", "cm2", "std", "sppa", "geos", "lat", "dismap")
setwd("~/GitHub/GIS") # move to download folder
for (i in chapters) {
  fn <- paste(i, "mod.R", sep = "_")
  download.file(paste(ASDAR_BOOK, fn, sep = "/"), fn)
}
list.files()
```
