---
title: "sawr11"
author: "Robert A. Stevens"
date: "January 4, 2017"
output: html_document
---

```{r, comment=NA}
library(maptools)
library(spdep)
library(DCluster)
library(R2WinBUGS)
library(rgdal)
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 11 Disease Mapping

## 11.1 Introduction

```{r, comment=NA}
nc_file <- system.file("shapes/sids.shp", package = "maptools")[1] 
llCRS <- CRS("+proj=longlat +datum=NAD27")
nc <- readShapePoly(nc_file, ID = "FIPSNO", proj4string = llCRS) 
rn <- sapply(slot(nc, "polygons"), function(x) slot(x, "ID"))
gal_file <- system.file("etc/weights/ncCR85.gal", package = "spdep")[1]
ncCR85 <- read.gal(gal_file, region.id = rn)
```

## 11.2 Statistical Models

```{r, comment=NA}
nc$Observed <- nc$SID74
nc$Population <- nc$BIR74
r <- sum(nc$Observed)/sum(nc$Population)
nc$Expected <- nc$Population*r
nc$SMR <- nc$Observed/nc$Expected
```

**Figure 11.1**. Standardized mortality ratio of the North Carolina SIDS data in the period 1974–1978

**Figure 11.2**. Confidence intervals of the SMR obtained with an exact formula. The black dot represents the SMR of each area. The confidence intervals shown by dashed lines are significantly higher than 1

### 11.2.1 Poisson-Gamma Model

```{r, comment=NA}
eb <- empbaysmooth(nc$Observed, nc$Expected) 
nc$EBPG <- eb$smthrr
```

**Figure 11.3**. Probability maps for the Poisson and negative binomial models

### 11.2.2 Log-Normal Model

```{r, comment=NA}
ebln <- lognormalEB(nc$Observed, nc$Expected) 
nc$EBLN <- exp(ebln$smthrr)
```

### 11.2.3 Marshall’s Global EB Estimator

```{r, comment=NA}
EBMarshall <- EBest(nc$Observed, nc$Expected) 
nc$EBMarshall <- EBMarshall[, 2]
```

**Figure 11.4**. Comparison of different risk estimators. SMR displays the standardized mortality ratio, whilst EBPG, EBLN, and EBMarshall show different empirical Bayes estimates using the Poisson-Gamma model, the log-normal model, and Marshall’s global estimator

## 11.3 Spatially Structured Statistical Models

```{r, comment=NA}
nc$EBMrshloc <- EBlocal(nc$Observed, nc$Expected, ncCR85)$est
```

**Figure 11.5**. Marshall’s EB estimator using local (top) and global (bottom) information

**Figure 11.6**. Comparison of raw and EB estimators of the relative risk

## 11.4 Bayesian Hierarchical Models

### 11.4.1 The Poisson-Gamma Model Revisited

```{r, comment=NA}
model {
  for(i in 1:N) {
    observed[i] ~ dpois(mu[i])
    mu[i] <- theta[i]*expected[i]
    theta[i] ~ dgamma(nu, alpha)
  }
  nu ~ dgamma(0.01, 0.01)
  alpha ~ dgamma(0.01, 0.01)          
}
```

**Figure 11.7**. Code of the Poisson-Gamma model for WinBugs

```{r, comment=NA}
N <- length(nc$Observed)
d <- list(N = N, observed = nc$Observed, expected = nc$Expected)
pgmodelfile <- paste(getwd(), "/PG-model.txt", sep = "")
wdir <- paste(getwd(), "/PG", sep = "") 
if (!file.exists(wdir)) {
  dir.create(wdir)
}
BugsDir <- "/home/asdar/.wine/dosdevices/c:/Program Files/WinBUGS14" 
MCMCres <- bugs(data = d, 
                inits = list(list(nu = 1, alpha = 1)), 
                working.directory = wdir, 
                parameters.to.save = c("theta", "nu", "alpha"), 
                n.chains = 1, 
                n.iter = 20000, 
                n.burnin = 10000, 
                n.thin = 10, 
                model.file = pgmodelfile, 
                bugs.directory = BugsDir, 
                WINEPATH = "/usr/bin/winepath")
```

```{r, comment=NA}
nc$PGmean <- MCMCres$mean$theta
nc$PGmedian <- MCMCres$median$theta
```

**Figure 11.8**. 95% credible intervals of the relative risks obtained with WinBUGS using a full Bayes Poisson-Gamma model

**Figure 11.9**. Comparison of empirical Bayes and full Bayes estimates of the relative risks using a Poisson-Gamma model

### 11.4.2 Spatial Models

```{r, comment=NA}
nc.nb <- nb2WB(ncCR85)
```

**Figure 11.10**. Proportion of non-white births in Carolina, 1974–1978. Notice the similar pattern to the relative risk estimates

```{r, comment=NA}
model {
  for(i in 1:N) {
    observed[i] ~ dpois(mu[i])
    log(theta[i]) <- alpha + beta*nonwhite[i] + u[i] + v[i] 
    mu[i] <- expected[i]*theta[i]
    u[i] ~ dnorm(0, precu)
  }
  
  v[1:N] ~ car.normal(adj[], weights[], num[], precv)
  
  alpha ~ dflat()
  beta ~ dnorm(0, 1.0E-5)
  precu ~ dgamma(0.001, 0.001)
  precv ~ dgamma(0.1, 0.1)
  sigmau<-1/precu
  sigmav<-1/precv
}
```

**Figure 11.11**. Code of the Besag-York-Molli ́e model for WinBugs

```{r, comment=NA}
nc$nwprop <- nc$NWBIR74/nc$BIR74
d <- list(N = N, 
          observed = nc$Observed, 
          expected = nc$Expected, 
          nonwhite = nc$nwprop, 
          adj = nc.nb$adj, 
          weights = nc.nb$weights,
          num = nc.nb$num)
dwoutcov <- list(N = N, 
                 observed = nc$Observed, 
                 expected = nc$Expected, 
                 adj = nc.nb$adj, 
                 weights = nc.nb$weights, 
                 num = nc.nb$num)
inits <- list(u = rep(0, N), 
              v = rep(0, N), 
              alpha = 0, 
              beta = 0, 
              precu = 0.001, 
              precv = 0.001)
```

```{r, comment=NA}
bymmodelfile <- paste(getwd(), "/BYM-model.txt", sep = "")
wdir <- paste(getwd(), "/BYM", sep = "") 
if (!file.exists(wdir)) {
 dir.create(wdir)
}
BugsDir <- "/home/asdar/.wine/dosdevices/c:/Program Files/WinBUGS14"
MCMCres <- bugs(data = d, 
                inits = list(inits), 
                working.directory = wdir, 
                parameters.to.save = c("theta", "alpha", "beta", "u", 
                                       "v", "sigmau", "sigmav"), 
                n.chains = 1, 
                n.iter = 30000, 
                n.burnin = 20000,
                n.thin = 10, 
                model.file = bymmodelfile, 
                bugs.directory = BugsDir, 
                WINEPATH = "/usr/bin/winepath")
```

```{r, comment=NA}
nc$BYMmean <- MCMCres$mean$theta
nc$BYMumean <- MCMCres$mean$u
nc$BYMvmean <- MCMCres$mean$v
```

**Figure 11.12**. Plots of the posterior distributions of α, β, and the deviance of the model

```{r, comment=NA}
geweke.diag(ncoutput[, c("deviance", "alpha", "beta", "theta[94]")])
```

**Figure 11.13**. Standardized Mortality Ratio and posterior means of the relative risks obtained with the BYM model

**Figure 11.14**. Posterior means of the non-spatial random effects (u[i]) estimated with the BYM model

**Figure 11.15**. Posterior means of the spatial random effects (v[i]) estimated with the BYM model

**Figure 11.16**. 95% credible intervals of the relatives risks obtained with the BYM model

## 11.5 Detection of Clusters of Disease

### 11.5.1 Testing the Homogeneity of the Relative Risks

```{r, comment=NA}
chtest <- achisq.test(Observed ~ offset(log(Expected)), 
                      as(nc, "data.frame"), 
                      "multinom", 
                      999)
chtest
```

```{r, comment=NA}
1 - pchisq(chtest$t0, 100 - 1)
```

```{r, comment=NA}
pwtest <- pottwhitt.test(Observed ~ offset(log(Expected)), 
                         as(nc, "data.frame"), 
                         "multinom", 
                         999)
```

```{r, comment=NA}
Oplus <- sum(nc$Observed)
1 - pnorm(pwtest$t0, Oplus*(Oplus - 1), sqrt(2*100*Oplus*(Oplus - 1)))
```

### 11.5.2 Moran’s I Test of Spatial Autocorrelation

```{r, comment=NA}
col.W <- nb2listw(ncCR85, zero.policy = TRUE)
moranI.test(Observed ~ offset(log(Expected)), 
            as(nc, "data.frame"), 
            "negbin", 
            999, 
            listw = col.W, 
            n = length(ncCR85), 
            S0 = Szero(col.W))
```

### 11.5.3 Tango’s Test of General Clustering

```{r, comment=NA}
data(nc.sids)
idx <- match(nc$NAME, rownames(nc.sids))
nc$x <- nc.sids$x[idx]
nc$y <- nc.sids$y[idx]
coords <- cbind(nc$x, nc$y)
dlist <- dnearneigh(coords, 0, Inf)
dlist <- include.self(dlist)
dlist.d <- nbdists(dlist, coords)
phi <- 100
col.W.tango <- nb2listw(dlist, 
                        glist = lapply(dlist.d, function(x, phi) {exp(-x/phi)}, phi = phi), 
                        style = "C")
```

```{r, comment=NA}
tango.test(Observed ~ offset(log(Expected)), 
           as(nc, "data.frame"), 
           "negbin", 
           999, 
           listw = col.W.tango, 
           zero.policy = TRUE)
```

### 11.5.4 Detection of the Location of a Cluster

### 11.5.5 Geographical Analysis Machine

```{r, comment=NA}
sidsgam <- opgam(data = as(nc, "data.frame"), radius = 30, step = 10, alpha = 0.002)
gampoints <- SpatialPoints(sidsgam[ , c("x", "y")]*1000, 
                           CRS("+proj=utm +zone=18 +datum=NAD27"))
ll <- CRS("+proj=longlat +datum=NAD27")
gampoints <- spTransform(gampoints, ll)
gam.layout <- list("sp.points", gampoints)
```

**Figure 11.17**. Results of Openshaw’s GAM. The dots represent the centre of the clusters

### 11.5.6 Kulldorff’s Statistic

```{r, comment=NA}
mle <- calculate.mle(as(nc, "data.frame"), model = "negbin") 
thegrid <- as(nc, "data.frame")[ , c("x", "y")]
knresults <- opgam(data = as(nc, "data.frame"), 
                   thegrid = thegrid, 
                   alpha = 0.05, 
                   iscluster = kn.iscluster, 
                   fractpop = 0.15, 
                   R = 99, 
                   model = "negbin", 
                   mle = mle)
```

**Figure 11.18**. Results of Kulldorff’s test. The circles show the most likely cluster

### 11.5.7 Stone’s Test for Localized Clusters

```{r, comment=NA}
stone.stat(as(nc, "data.frame"), region = which(nc$NAME == "Anson"))
st <- stone.test(Observed ~ offset(log(Expected)), 
                 as(nc, "data.frame"), 
                 model = "negbin", 
                 99, 
                 region = which(nc$NAME == "Anson"))
st
```

## 11.6 Other Topics in Disease Mapping
