---
title: "sawr10"
author: "Robert A. Stevens"
date: "January 4, 2017"
output: html_document
---

```{r, comment=NA}
library(spdep)
library(nlme)
library(lmtest)
library(sandwich)
library(mgcv)
library(MASS)
library(spgwr)
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 10 Modeling Areal Data

## 10.1 Introduction

## 10.2 Spatial Statistics Approaches

```{r, comment=NA}
nylm <- lm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8) 
summary(nylm)
NY8$lmresid <- residuals(nylm)
```

**Figure 10.1**. Residuals from the linear model of transformed incidence proportions; TCE site locations shown for comparative purposes

```{r, comment=NA}
NYlistw <- nb2listw(NY_nb, style = "B") 
lm.morantest(nylm, NYlistw)
```

### 10.2.1 Simultaneous Autoregressive Models

```{r, comment=NA}
nysar <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, listw = NYlistw)
summary(nysar)
```

```{r, comment=NA}
nylmw <- lm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, weights = POP8)
summary(nylmw)
NY8$lmwresid <- residuals(nylmw)
```

```{r, comment=NA}
lm.morantest(nylmw, NYlistw)
```

**Figure 10.2**. Residuals from the weighted linear model of transformed incidence proportions; TCE site locations shown for comparative purposes

```{r, comment=NA}
nysarw <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, listw = NYlistw, weights = POP8)
summary(nysarw)
```

### 10.2.2 Conditional Autoregressive Models

```{r, comment=NA}
nycar <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                  data = NY8, family = "CAR", listw = NYlistw)
summary(nycar)
```

```{r, comment=NA}
nycarw <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                   data = NY8, family = "CAR", listw = NYlistw, weights = POP8) 
summary(nycarw)
```

### 10.2.3 Fitting Spatial Regression Models

```{r, comment=NA}
nysarwM <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                    data = NY8, family = "SAR", listw = NYlistw, weights = POP8, 
                    method = "Matrix")
summary(nysarwM)
```

**Figure 10.3**. Log likelihood values for a range of values of λ, weighted and unweighted SAR models; fitted spatial coefficient values and maxima shown

```{r, comment=NA}
1/range(eigenw(NYlistw))
nysar_ll  <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                      data = NY8, family = "SAR", listw = NYlistw, llprof = 100)
nysarw_ll <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                      data = NY8, family = "SAR", listw = NYlistw, llprof = 100,
                      weights = POP8)
```

```{r, comment=NA}
nysmaw <- spautolm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                   data = NY8, family = "SMA", listw = NYlistw, weights = POP8) 
summary(nysmaw)
```

## 10.3 Mixed-Effects Models

```{r, comment=NA}
NY8$x <- coordinates(NY8)[ , 1]/1000
NY8$y <- coordinates(NY8)[ , 2]/1000
sp1 <- corSpatial(1, form = ~x + y, type = "gaussian") 
scor <- Initialize(sp1, as(NY8, "data.frame")[ , c("x", "y")], nugget = FALSE)
```

```{r, comment=NA}
spmodel <- lme(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
               random = ~1 | AREAKEY, 
               data = as(NY8, "data.frame"), correlation = scor, method = "ML")
summary(spmodel)
```

## 10.4 Spatial Econometrics Approaches

```{r, comment=NA}
bptest(nylm)
```

```{r, comment=NA}
coeftest(nylm)
coeftest(nylm, vcov = vcovHC(nylm, type = "HC4"))
```

```{r, comment=NA}
NYlistwW <- nb2listw(NY_nb, style = "W")
res <- lm.LMtests(nylm, listw = NYlistwW, test = "all")
tres <- t(sapply(res, function(x) c(x$statistic, x$parameter, x$p.value)))
colnames(tres) <- c("Statistic", "df", "p-value") 
printCoefmat(tres)
```

```{r, comment=NA}
nylag <- lagsarlm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, listw = NYlistwW)
summary(nylag)
bptest.sarlm(nylag)
```

```{r, comment=NA}
nymix <- lagsarlm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                  data = NY8, listw = NYlistwW, type = "mixed")
nymix
anova(nymix, nylag)
```

```{r, comment=NA}
nyerr <- errorsarlm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, listw = NYlistwW)
summary(nyerr)
```

```{r, comment=NA}
nystsls <- stsls(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY8, listw = NYlistwW)
summary(nystsls)
```

```{r, comment=NA}
nystslsR <- stsls(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                  data = NY8, listw = NYlistwW, robust = TRUE)
summary(nystslsR)
```

```{r, comment=NA}
nyGMerr <- GMerrorsar(Z ~ PEXPOSURE, data = NY8, listw = NYlistwW)
summary(nyGMerr)
```

## 10.5 Other Methods

### 10.5.1 GAM, GEE, GLMM

```{r, comment=NA}
nyGAM1 <- gam(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + s(x, y), weights = POP8, data = NY8)
anova(nylmw, nyGAM1, test = "Chisq")
```

```{r, comment=NA}
nyGLMp <- glm(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)), 
              data = NY8, family = "poisson")
summary(nyGLMp)
```

```{r, comment=NA}
NY8$lmpresid <- residuals(nyGLMp, type = "deviance") 
lm.morantest(nyGLMp, listw = NYlistwW)
```

**Figure 10.5**. Residuals from the Poisson regression model; TCE site locations shown for comparative purposes

```{r, comment=NA}
nyGAMp <- gam(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)) + s(x, y), 
              data = NY8, family = "poisson") 
summary(nyGAMp)
anova(nyGLMp, nyGAMp, test = "Chisq")
```

```{r, comment=NA}
attach(as(NY8, "data.frame"))
nyGLMMp <- glmmPQL(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)), 
                   data = NY8, family = poisson, random = ~1 | AREAKEY, correlation = scor)
detach("as(NY8, \"data.frame\")")
summary(nyGLMMp)
```

### 10.5.2 Moran Eigenvectors

```{r, comment=NA}
nySFE <- SpatialFiltering(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
                          data = NY8, nb = NY_nb, style = "W", verbose = FALSE)
nylmSFE <- lm(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + fitted(nySFE), data = NY8)
summary(nylmSFE)
anova(nylm, nylmSFE)
```

```{r, comment=NA}
nyME <- ME(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
           data = NY8, offset = log(POP8), family = "poisson", listw = NYlistwW, alpha = 0.5)
nyME
nyglmME <- glm(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)) + fitted(nyME),
               data = NY8, family = "poisson")
summary(nyglmME)
anova(nyGLMp, nyglmME, test = "Chisq")
```

**Figure 10.6**. Maps of the two eigenvalues selected for inclusion in the Poisson regression model

### 10.5.3 Geographically Weighted Regression

```{r, comment=NA}
bwG <- gwr.sel(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
               data = NY8, gweight = gwr.Gauss, verbose = FALSE)
gwrG <- gwr(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, 
            data = NY8, bandwidth = bwG, gweight = gwr.Gauss, hatmatrix = TRUE)
gwrG
```

```{r, comment=NA}
gbwG <- ggwr.sel(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)), 
                 data = NY8, family = "poisson", gweight = gwr.Gauss, verbose = FALSE)
ggwrG <- ggwr(Cases ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME + offset(log(POP8)), 
              data = NY8, family = "poisson", bandwidth = gbwG, gweight = gwr.Gauss)
ggwrG
```

**Figure 10.7**. GWR local coefficient estimates for the exposure to TCE site covariate

**Figure 10.8**. Pairs plots of GWR local coefficient estimates showing the effects of GWR collinearity forcing
