---
title: "sawr03"
author: "Robert A. Stevens"
date: "December 5, 2016"
output: html_document
---

```{r, comment=NA}
library(sp)
library(maptools)
library(maps)
library(rgdal)
library(lattice)
library(grid)
library(RColorBrewer)
library(classInt)
library(gstat)
library(ggplot2)
library(latticeExtra)
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 3 Visualizing Spatial Data

## 3.1 The Traditional Plot System

### 3.1.1 Plotting Points, Lines, Polygons, and Grids

```{r, comment=NA}
data(meuse)
coordinates(meuse) <- c("x", "y")
plot(meuse)
title("points")
```

**Figure 3.1 (a)**. The meuse data set: sample points

```{r, comment=NA}
cc <- coordinates(meuse)
m.sl <- SpatialLines(list(Lines(list(Line(cc)), "line1")))
plot(m.sl)
title("lines")
```

**Figure 3.1 (b)**. The meuse data set: the sample path (line)

```{r, comment=NA}
data(meuse.riv)
meuse.lst <- list(Polygons(list(Polygon(meuse.riv)), "meuse.riv"))
meuse.sr <- SpatialPolygons(meuse.lst)
plot(meuse.sr, col = "grey")
title("polygons")
```

**Figure 3.1 (c)**. The meuse data set: the Meuse river (ring)

```{r, comment=NA}
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
meuse.grid <- as(meuse.grid, "SpatialPixels") 
image(meuse.grid, col = "grey")
title("grid")
```

**Figure 3.1 (d)**. The meuse data set: the gridded study area

```{r, comment=NA}
image(meuse.grid, col = "khaki2")
#plot(meuse.pol, col = "lightsteelblue2", add = TRUE)
plot(meuse.sr, col = "lightsteelblue2", add = TRUE)
plot(meuse, add = TRUE, col = "brown", cex = 0.5)
```

**Figure 3.2**. Map elements combined into a single map

### 3.1.2 Axes and Layout Elements

```{r, comment=NA}
layout(matrix(c(1, 2), 1, 2))
plot(meuse.sr, axes = TRUE)
plot(meuse.sr, axes = FALSE)
axis(1, at = c(178000 + 0:2*2000), cex.axis = 0.7) 
axis(2, at = c(326000 + 0:3*4000), cex.axis = 0.7) 
box()
```

**Figure 3.3**. Default axes (left) and custom axes (right) for the meuse.riv data

**Table 3.1**. Graphic arguments useful for controlling figure and plotting region

Argument Meaning          Unit          Length  
-------- ---------------- ------------- ------  
fin      Figure region    Inch          2  
pin      Plotting region  Inch          2  
mai      Plotting margins Inch          4  
mar      Plotting margins Lines of text 4  

see ?par for more information

```{r, comment=NA}
oldpar = par(no.readonly = TRUE)
layout(matrix(c(1, 2), 1, 2))
plot(meuse, axes = TRUE, cex = 0.6) 
plot(meuse.sr, add = TRUE)
title("Sample locations")
par(mar = c(0, 0, 0, 0) + 0.1)
plot(meuse, axes = FALSE, cex = 0.6)
plot(meuse.sr, add = TRUE)
box()
par(oldpar)
```

**Figure 3.4**. Equal-area plots with (left) and without (right) the default space R reserves for axes and title(s)

```{r, comment=NA}
plot(meuse)
#plot(meuse.pol, add = TRUE)
plot(meuse.sr, add = TRUE)
#plot(meuse)
SpatialPolygonsRescale(layout.scale.bar(), offset = c(180200, 329600),
    scale = 1000, fill = c("transparent", "black"), plot.grid = FALSE)
text(x = c(180200, 181200), y = rep(329750, 2), c("0", "1 km"))
SpatialPolygonsRescale(layout.north.arrow(), offset = c(178750, 332500),
    scale = 400, plot.grid = FALSE)
box()
```

**Figure 3.5**. Scale bar and north arrow as map elements

### 3.1.3 Degrees in Axes Labels and Reference Grid

```{r, comment=NA}
nc <- readShapePoly(system.file("shapes/sids.shp", package = "maptools")[1], 
                    proj4string = CRS("+proj=longlat +datum=NAD27"))
rrt <- nc$SID74/nc$BIR74
brks <- quantile(rrt, seq(0, 1, 1/5))
cols <- brewer.pal(5, "Reds")
plot(nc, col = cols[findInterval(rrt, brks, all.inside = TRUE)], axes = FALSE)
box()
degAxis(1)
degAxis(2, at=34:37)
```

**Figure 3.6**. Decimal degrees in axis labels: the North Carolina SIDS data

### 3.1.4 Plot Size, Plotting Area, Map Scale, and Multiple Plots

```{r, comment=NA, eval=FALSE}
par("pin")
par(pin = c(4, 4))
```

```{r, comment=NA, eval=FALSE}
dev.off()
X11(width = 10, height = 10)
```

```{r, comment=NA, eval=FALSE}
postscript("file.ps", width = 10, height = 10)
```

```{r, comment=NA}
pin <- par("pin")
dxy <- apply(bbox(meuse), 1, diff)
ratio <- dxy[1]/dxy[2]
par(pin = c(ratio*pin[2], pin[2]), xaxs = "i", yaxs = "i") 
plot(meuse, pch = 1)
box()
```

**Figure 3.7**. Plotting region exactly equal to sample location ranges: border point symbols are clipped

```{r, comment=NA, eval=FALSE}
par(mfrow = c(2, 3))
```

```{r, comment=NA, eval=FALSE}
layout(matrix(1:6, 2, 3, byrow = TRUE))
```

### 3.1.5 Plotting Attributes and Map Legends

**Table 3.2**. Useful annotation arguments to be passed to plot or image methods

- **Class(Further help)**
    + **Argument: Meaning**

- SpatialLinesDataFrame (?lines)
    + col: Color 
    + lwd: Line width 
    + lty: Line type

- SpatialPolygonsDataFrame (?polygon)
    + border: Border color 
    + density: Hashing density 
    + angle: Hashing angle 
    + lty: Line type
    + pbg: Hole color

- SpatialPointsDataFrame (?points)
    + pch: Symbol
    + col: Color
    + bg: Fill color 
    + cex: Symbol size

- SpatialPixelsDataFrame [a] and SpatialGridDataFrame (?image.default)
    + zlim: Attribute value limits 
    + col: Colors
    + breaks: Break points

[a] Use image to plot gridded data

```{r, comment=NA}
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
gridded(meuse.grid) <- TRUE
zn.idw <- krige(log(zinc) ~ 1, meuse, meuse.grid)

cols <- brewer.pal(4, "Reds")
image(zn.idw, col = cols, breaks = log(c(100, 200, 400, 800, 1800)))
#plot(meuse.pol, add = TRUE)
plot(meuse.sr, add = TRUE)
plot(meuse, pch = 1, cex = sqrt(meuse$zinc)/20, add = TRUE)
legVals <- c(100, 200, 500, 1000, 2000)
legend("left", legend = legVals, pch = 1, pt.cex = sqrt(legVals)/20, bty = "n",
       title = "measured, ppm", cex = 0.8, y.inter = 0.9)
legend("topleft", fill = cols, legend = c("100-200", "200-400", "400-800", "800-1800"), 
       bty = "n", title = "interpolated, ppm", cex = 0.8, y.inter = 0.9)
title("measured and interpolated zinc")
```

**Figure 3.8**. Sample data points for zinc (ppm) plotted over an interpolated image, with symbol area proportional to measured concentration

## 3.2 Trellis/Lattice Plots with spplot

### 3.2.1 A Straight Trellis Example

```{r, comment=NA}
data(meuse)
coordinates(meuse) <- ~x + y
data(meuse.grid)
coordinates(meuse.grid) <- ~x + y
gridded(meuse.grid) <- TRUE
zn <- krige(zinc ~ 1, meuse, meuse.grid)
zn$direct <- zn$var1.pred
zn$log <- exp(krige(log(zinc) ~ 1, meuse, meuse.grid)$var1.pred)

levelplot(z ~ x + y | name, spmap.to.lev(zn[c("direct", "log")]), asp = "iso")
spplot(zn[c("direct", "log")])
```

**Figure 3.9**. Two interpolation scenarios for the meuse data set, plotted on the same total size. (Top) Example of levelplot, (bottom) example of the spplot wrapper, which turns off axes

### 3.2.2 Plotting Points, Lines, Polygons, and Grids

```{r, comment=NA}
cuts <- c(0, 20, 50, 200, 500, 2000)
grys <- brewer.pal(7, "Reds")
print(spplot(meuse[1:4], main = "ppm", cuts = cuts, cex = 0.5, col.regions = grys, 
             key.space = "right"),
      split = c(1, 1, 1, 2), more = TRUE)

meuse$lead.st <- as.vector(scale(meuse$lead))
meuse$zinc.st <- as.vector(scale(meuse$zinc))
meuse$copper.st <- as.vector(scale(meuse$copper))
meuse$cadmium.st <- as.vector(scale(meuse$cadmium))
cuts <- c(-1.2, 0, 1, 2, 3, 5)
print(spplot(meuse, c("cadmium.st", "copper.st", "lead.st", "zinc.st"), key.space = "right", 
             main = "standardised", cex = 0.5, cuts = cuts, col.regions = grys),
      split=c(1, 2, 1, 2))
```

**Figure 3.10**. Soil measurements for four heavy metals in the Meuse data set; (top) in ppm units, (bottom) each variable scaled to mean zero and unit standard variance

```{r, comment=NA}
grys <- brewer.pal(9, "Reds")
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
meuse.grid <- as(meuse.grid, "SpatialPixelsDataFrame")
cl <- ContourLines2SLDF(contourLines(as.image.SpatialGridDataFrame(meuse.grid["dist"])))
print(spplot(cl, colorkey = list(height = 0.8, width = 0.6), col.regions = grys), 
      split = c(1, 1, 3, 1), more = TRUE)
grys <- brewer.pal(6, "Reds")
cuts <- (0:5)/5
print(spplot(meuse.grid, "dist", 
             at = cuts, 
             colorkey = list(labels = list(at = cuts), at = cuts),
             col.regions = grys), 
      split = c(2, 1, 3, 1), more = TRUE)
meuse.grid$f <- factor(meuse.grid$ffreq, labels = c("annual", "2-5 yrs", "> 5 yrs"))
print(spplot(meuse.grid, "f", colorkey = list(height = 0.4, width = 0.6), 
             col.regions = brewer.pal(3, "Reds")), split = c(3, 1, 3, 1), more = FALSE)
```

**Figure 3.11**. (Left) Contour lines for distance to river Meuse, levels represented by grey tones; (middle) grid plot of a numerical variable; (right) plot of the factor variable flood frequency; note the different legend key

### 3.2.3 Adding Reference and Layout Elements to Plots

sp layout function Object class    Useful arguments [a]  
------------------ --------------- --------------------  
sp.points          SpatialPoints   pch, cex, col  
sp.polygons        SpatialPolygons lty, lwd, col  
sp.lines           SpatialLines    lty, lwd, col  
sp.text            text            (see panel.text)  

[a] For help, see ?par


```{r, comment=NA}
river <- list("sp.polygons", meuse.sr)
north <- list("SpatialPolygonsRescale", layout.north.arrow(), 
              offset = c(178750, 332500), scale = 400)
scale <- list("SpatialPolygonsRescale", layout.scale.bar(), 
              offset = c(180200, 329800), scale = 1000,
              fill = c("transparent", "black"))
txt1 <- list("sp.text", c(180200, 329950), "0")
txt2 <- list("sp.text", c(181200, 329950), "1 km")
pts <- list("sp.points", meuse, pch = 3, col = "black") 
meuse.layout <- list(river, north, scale, txt1, txt2, pts)
spplot(zn["log"], sp.layout = meuse.layout)
```

**Figure 3.12**. Interpolated spplot image with layout elements

### 3.2.4 Arranging Panel Layout

## 3.3 Interacting with Plots

### 3.3.1 Interacting with Base Graphics

```{r, comment=NA, eval=FALSE}
plot(meuse)
meuse.id <- identify(coordinates(meuse))
```

```{r, comment=NA}
# plot(meuse)
# region <- locator(type = "o")
# n <- length(region$x)
# p <- Polygon(cbind(region$x, region$y)[c(1:n, 1), ], hole = FALSE)
# ps <- Polygons(list(p), ID = "region")
# sps <- SpatialPolygons(list(ps))
# plot(meuse[!is.na(overlay(meuse, sps)), ], pch = 16, cex = 0.5, add = TRUE)
prj <- CRS("+proj=longlat +datum=NAD27")
nc_shp <- system.file("shapes/sids.shp", package = "maptools")[1] 
nc <- readShapePoly(nc_shp, proj4string = prj)
# plot(nc)
# pt <- locator(type = "p")
# print(pt)
# overlay(nc, SpatialPoints(cbind(pt$x, pt$y), proj4string = prj))
```

**Figure 3.13**. Interaction with point plots. (Left) Individual identification of points; (right) digitizing a region, highlighted points included in the region

### 3.3.2 Interacting with spplot and Lattice Plots

```{r, comment=NA, eval=FALSE}
ids <- spplot(meuse, "zinc", identify = TRUE)
```

```{r, comment=NA, eval=FALSE}
trellis.focus("panel", column = 1, row = 1) 
ids <- panel.identify()
trellis.unfocus()
```

```{r, comment=NA, eval=FALSE}
trellis.focus("panel", column = 1, row = 1) 
as.numeric(grid.locator())
trellis.unfocus()
```

## 3.4 Color Palettes and Class Intervals

### 3.4.1 Color Palettes

```{r, comment=NA}
rw.colors <- colorRampPalette(c("red", "white")) 
image(meuse.grid["dist"], col = rw.colors(10))
```

```{r, comment=NA, eval=FALSE}
example(brewer.pal)
```

### 3.4.2 Class Intervals

```{r, comment=NA}
pal <- brewer.pal(5, "Reds")
q5 <- classIntervals(meuse$zinc, n = 5, style = "quantile")
q5
diff(q5$brks)
plot(q5, pal = pal)
```

**Figure 3.14**. Comparison of quantile and natural breaks methods for setting class intervals, Meuse bank zinc ppm

```{r, comment=NA}
fj5 <- classIntervals(meuse$zinc, n = 5, style = "fisher") 
fj5
diff(fj5$brks)
plot(fj5, pal = pal)
```

```{r, comment=NA}
oopar <- par(mar = c(1, 1, 3, 1) + 0.1, mfrow=c(1, 2))
q5Colours <- findColours(q5, pal)
plot(meuse, col = q5Colours, pch = 19)
points(meuse, pch = 1)
box()
title(main = "Quantile")
legend("topleft", fill = attr(q5Colours, "palette"), legend = names(attr(q5Colours, "table")),
       bty = "n", cex = 0.8, y.intersp = 0.8)
fj5Colours <- findColours(fj5, pal)
plot(meuse, col = fj5Colours, pch = 19)
points(meuse, pch = 1)
box()
title(main = "Fisher-Jenks")
legend("topleft", fill = attr(fj5Colours, "palette"), legend = names(attr(fj5Colours, "table")),
       bty = "n", cex = 0.8, y.intersp = 0.8)
par(oopar)
```

**Figure 3.15**. Comparison of output maps made with quantile and natural breaks class intervals, Meuse bank zinc ppm

```{r, comment=NA}
cuts = (0:10)/10
spplot(meuse.grid, "dist", colorkey = list(labels = list(at = cuts)), at = cuts)
```
