---
title: "sawr04"
author: "Robert A. Stevens"
date: "December 20, 2016"
output: html_document
---

```{r, comment=NA}
library(rgdal)
library(maptools)
library(spdep)
library(DCluster)
library(RColorBrewer)
library(rgeos)
library(spacetime)
library(gstat)
library(osmar)
library(spgrass6)
#library(RgoogleMaps)
library(geoR) # Analysis of Geostatistical Data
# For an Introduction to geoR go to http://www.leg.ufpr.br/geoR
setwd("~/GitHub/GIS")
```

*Applied Spatial Data Analysis with R* by Roger S. Bivand, Edzer J. Pebesma, and Virgilio Gómez-Rubio

# 4 Spatial Data Import and Export

```{r, comment=NA, eval=FALSE}
library(rgdal)
```

## 4.1 Coordinate Reference Systems

### 4.1.1 Using the EPSG List

```{r, comment=NA}
EPSG <- make_EPSG()
EPSG[grep("^# ED50$", EPSG$note), ]
```

### 4.1.2 PROJ.4 CRS Specification

```{r, comment=NA}
CRS("+init=epsg:4230")
```

```{r, comment=NA}
ED50 <- CRS("+init=epsg:4230 +towgs84=-87,-96,-120,0,0,0,0")
ED50
```

### 4.1.3 Projection and Transformation

```{r, comment=NA}
IJ.east <- as(char2dms("4d31'00\"E"), "numeric")
IJ.north <- as(char2dms("52d28'00\"N"), "numeric")
IJ.ED50 <- SpatialPoints(cbind(x = IJ.east, y = IJ.north), ED50)
res <- spTransform(IJ.ED50, CRS("+proj=longlat +datum=WGS84"))
x <- as(dd2dms(coordinates(res)[1]), "character")
y <- as(dd2dms(coordinates(res)[2], TRUE), "character") 
cat(x, y, "\n")
spDistsN1(coordinates(IJ.ED50), coordinates(res), longlat = TRUE)*1000
gzAzimuth(coordinates(IJ.ED50), coordinates(res))
```

```{r, comment=NA}
proj4string(IJ.ED50) <- CRS("+init=epsg:4230")
res <- spTransform(IJ.ED50, CRS("+proj=longlat +datum=WGS84"))
spDistsN1(coordinates(IJ.ED50), coordinates(res), longlat=TRUE)*1000
gzAzimuth(coordinates(IJ.ED50), coordinates(res))
```

```{r, comment=NA}
EPSG[grep("Atlas", EPSG$note), 1:2]
CRS("+init=epsg:2163")
```

```{r, comment=NA}
proj <- projInfo("proj")
proj[proj$name == "laea", ]
ellps <- projInfo("ellps")
ellps[grep("a=6370997", ellps$major), ]
```

### 4.1.4 Degrees, Minutes, and Seconds

```{r, comment=NA}
IJ.dms.E <- "4d31'00\"E" 
IJ.dms.N <- "52d28'00\"N"
```

```{r, comment=NA}
IJ_east <- char2dms(IJ.dms.E)
IJ_north <- char2dms(IJ.dms.N)
IJ_east
IJ_north
getSlots("DMS")
```

```{r, comment=NA}
c(as(IJ_east, "numeric"), as(IJ_north, "numeric"))
```

## 4.2 Vector File Formats

### 4.2.1 Using OGR Drivers in rgdal

```{r, comment=NA}
head(ogrDrivers(), n = 10)

scot_dat <- read.table("scotland.dat", skip = 1)
names(scot_dat) <- c("District", "Observed", "Expected", "PcAFF", "Latitude", "Longitude")

ogrInfo(".", "scot")

scot_LL <- readOGR(dsn = ".", layer = "scot", integer64 = "allow.loss")
proj4string(scot_LL)
proj4string(scot_LL) <- CRS("+proj=longlat +ellps=WGS84")

sapply(slot(scot_LL, "data"), class)
scot_LL$ID
```

```{r, comment=NA}
#scot_dat <- read.table("scotland.dat", skip = 1) 
#names(scot_dat) <- c("District", "Observed", "Expected", "PcAFF", "Latitude", "Longitude")
#scot_dat$District
#scot_dat1 <- scot_dat[match(scot_LL$ID, scot_dat$District), ]
#row.names(scot_dat1) <- sapply(slot(scot_LL, "polygons"), function(x) slot(x, "ID"))
#scot_LLa <- spCbind(scot_LL, scot_dat1)
#all.equal(scot_LLa$ID, scot_LLa$District)
#names(scot_LLa)
#
scot_dat$District
ID_D <- match(scot_LL$ID, scot_dat$District)
scot_dat1 <- scot_dat[ID_D,]
row.names(scot_dat1) <- row.names(scot_LL)
scot_LLa <- spCbind(scot_LL, scot_dat1)
all.equal(scot_LLa$ID, scot_LLa$District)
names(scot_LLa)
```

```{r, comment=NA}
# O <- scot_LLa$Observed
# E <- scot_LLa$Expected
# scot_LLa$SMR <- probmap(O, E)$relRisk/100 
# scot_LLa$smth <- empbaysmooth(O, E)$smthrr
#
O <- scot_LLa$Observed
E <- scot_LLa$Expected
scot_LLa$SMR <- probmap(O, E)$relRisk/100
scot_LLa$smth <- empbaysmooth(O, E)$smthrr
```

```{r, comment=NA}
scot_BNG <- spTransform(scot_LLa, CRS("+init=epsg:27700"))

spplot(scot_BNG, c("SMR", "smth"),
 at = c(0, 0.25, 0.5, 0.8, 1, 1.5, 2.5, 4.5, 7),
 col.regions = rev(brewer.pal(8, "RdBu")))
```

**Figure 4.1**. Comparison of relative risk and EB smoothed relative risk for Scottish lip cancer

```{r, comment=NA}
# writeOGR(scot_LLa["ID"], dsn = "scot_district.kml",layer = "borders", driver = "KML")
# llCRS <- CRS("+proj=longlat ellps=WGS84")
# scot_SP_LL <- SpatialPointsDataFrame(
#   coordinates(scot_LLa), 
#   proj4string = llCRS, 
#   data = as(scot_LLa, "data.frame")[c("NAME", "Observed", "Expected", "SMR", "smth")])
# writeOGR(scot_SP_LL, dsn = "scot_rates.kml", layer = "rates", driver = "KML")
```

**Figure 4.2**. Scottish district boundaries and centroid pointers shown in Google Earth

```{r, comment=NA}
drv <- "ESRI Shapefile"
writeOGR(scot_BNG, dsn = ".", layer = "scot_BNG", driver = drv, overwrite_layer = TRUE)

list.files(pattern="^scot_BNG")
```

```{r, comment=NA}
load("geohub.RData")

#dsn <- "WFS:http://geohub.jrc.ec.europa.eu/effis/ows"
#ogrListLayers(dsn)

print(layers)

#Fires <- readOGR(dsn, "EFFIS:FiresAll")

geohub

Fires <- readOGR(".", "fires_120104")

names(Fires)

x <- c(-15, -15, 38, 38, -15)
y <- c(28, 62, 62, 28, 28)
crds <- cbind(x = x, y = y)
bb <- SpatialPolygons(list(Polygons(list(Polygon(coords=crds)), "1")))
data(wrld_simpl)
proj4string(bb) <- CRS(proj4string(wrld_simpl))
slbb <- gIntersection(bb, as(wrld_simpl, "SpatialLines"))
spl <- list("sp.lines", slbb, lwd = 0.7, col = "khaki4")

Fires$dt <- as.Date(as.character(Fires$FireDate), format = "%d-%m-%Y")
Fires0 <- Fires[-which(coordinates(Fires)[ , 2] < 0), ]
Fires1 <- Fires0[order(Fires0$dt), ]
Fires2 <- STIDF(as(Fires1, "SpatialPoints"), Fires1$dt, as(Fires1, "data.frame"))

stplot(as(Fires2, "STI"), number = 3, sp.layout = spl, cex = 0.5)

names(Fires1)[1] <- "name"
GR_Fires <- Fires1[Fires1$Country == "GR", ]
writeOGR(GR_Fires, "EFFIS.gpx", "waypoints", driver = "GPX", 
         dataset_options = "GPX_USE_EXTENSIONS=YES", overwrite_layer = TRUE, delete_dsn = TRUE)

GR <- readOGR("EFFIS.gpx", "waypoints")
GR[1, c(5, 24:28)]
```

### 4.2.2 Other Import/Export Functions

```{r, comment=NA}
getinfo.shape("scot_BNG.shp")
```

## 4.3 Raster File Formats

### 4.3.1 Using GDAL Drivers in rgdal

```{r, comment=NA}
auck_el1 <- readGDAL("70042108.tif")
summary(auck_el1)
is.na(auck_el1$band1) <- auck_el1$band1 <= 0 | auck_el1$band1 > 1e+4
```

```{r, comment=NA}
x <- GDAL.open("70042108.tif")
xx <- getDriver(x)
#xx # do not show pointer
getDriverLongName(xx)
#x # do not show pointer
dim(x)
GDAL.close(x)
```

```{r, comment=NA}
GDALinfo("70042108.tif")
#
brks <- c(0, 10, 20, 50, 100, 150, 200, 300, 400, 500, 600, 700)
pal <- terrain.colors(11)
pal
length(pal) == length(brks) - 1
auck_el1$band1 <- findInterval(auck_el1$band1, vec = brks, all.inside = TRUE) - 1
writeGDAL(auck_el1, "demIndex.tif", drivername = "GTiff", type = "Byte", 
          colorTable = list(pal), mvFlag = length(brks) - 1)
Gi <- GDALinfo("demIndex.tif", returnColorTable = TRUE)
CT <- attr(Gi, "ColorTable")[[1]]
CT[CT > "#000000"]
```

```{r, comment=NA}
data(meuse.grid)
coordinates(meuse.grid) <- c("x", "y")
gridded(meuse.grid) <- TRUE
proj4string(meuse.grid) <- CRS("+init=epsg:28992")
data(meuse)
coordinates(meuse) <- c("x", "y")
proj4string(meuse) <- CRS(proj4string(meuse.grid))

#log_zinc <- krige(log(zinc) ~ 1, meuse, meuse.grid)["var1.pred"]
log_zinc <- idw(log(zinc) ~ 1, meuse, meuse.grid)["var1.pred"]

summary(log_zinc)
writeGDAL(log_zinc, fname = "log_zinc.tif", drivername = "GTiff", type = "Float32",
          options = "INTERLEAVE=PIXEL")

GDALinfo("log_zinc.tif")
```

```{r, comment=NA}
Soil <- meuse.grid["soil"]
table(Soil$soil)
Soil$soil <- as.integer(Soil$soil) - 1
Cn <- c("Rd10A", "Rd90C/VII", "Bkd26/VII")
writeGDAL(Soil, "Soil.tif", drivername = "GTiff", type = "Byte", catNames = list(Cn),
          mvFlag = length(Cn))
Gi <- GDALinfo("Soil.tif", returnCategoryNames = TRUE)
attr(Gi, "CATlist")[[1]]
summary(readGDAL("Soil.tif"))

head(gdalDrivers(), n=10)

#service_xml <- "frmt_wms_openstreetmap_tms.xml"
#offset <- c(19339000, 34546000)
#osm <- readGDAL(service_xml, offset=offset, region.dim=c(2000, 2000), output.dim=c(1000, 1000))

load("getosm.RData")
cat(paste(strwrap(getosm, exdent = 5), collapse = "\n"), "\n")

osm <- readGDAL("osm_bergen_120105.tif")

summary(osm)

image(osm, red = 1, green = 2, blue = 2)

load("RgoogleMaps.RData")

#myMap <- GetMap(center=c(60.395, 5.322), zoom =16, destfile = "MyTile2.png", maptype = "mobile")

BB <- do.call("rbind", myMap$BBOX)
dBB <- rev(diff(BB))
DIM12 <- dim(myMap$myTile)[1:2]
cs <- dBB/DIM12
cc <- c(BB[1,2] + cs[1]/2, BB[1,1] + cs[2]/2)
GT <- GridTopology(cc, cs, DIM12)
p4s <- CRS("+proj=longlat +datum=WGS84")
SG_myMap <- SpatialGridDataFrame(
  GT, 
  proj4string = p4s, 
  data = data.frame(r = c(t(myMap$myTile[ , , 1]))*255, 
                    g = c(t(myMap$myTile[ , , 2]))*255, 
                    b = c(t(myMap$myTile[ , , 3]))*255))

#myMap1 <- GetMap.OSM(lonR = c(5.3190, 5.3280), latR = c(60.392, 60.398), scale = 4000, 
#                     destfile = "MyTile.png")

load("osmar.RData")

#api <- osmsource_api()
#box <- corner_bbox(5.3190, 60.392, 5.3280, 60.398)
#torget <- get_osm(box, source = api)

torget1 <- as_sp(torget, "lines")

sort(table(torget1$user), decreasing=TRUE)[1:3]

bybane <- find(torget, way(tags(k == "light_rail")))
bybane <- find_down(torget, way(bybane))
bybane <- subset(torget, ids=bybane)
bybane <- as_sp(bybane, "lines")

image(SG_myMap, red = 1, green = 2, blue = 3)
plot(torget1, add = TRUE)
plot(bybane, add = TRUE, lwd = 5, col = "orange2")
plot(0:1, 0:1, type = "n", axes = FALSE, asp = 1)
rasterImage(myMap1[[4]], 0, 0, 1, 1)

writeOGR(Fires[ , c("gml_id", "FireDate", "Area_HA")], 
         dsn = "fires.kml", 
         layer = "fires", 
         driver = "KML", 
         overwrite_layer = TRUE)
```

### 4.3.2 Writing a Google Earth Image Overlay

```{r, comment=NA}
grd <- as(meuse.grid, "SpatialPolygons")
proj4string(grd) <- CRS(proj4string(meuse))
grd.union <- unionSpatialPolygons(grd, rep("x", length(slot(grd, "polygons"))))
ll <- CRS("+proj=longlat +datum=WGS84")
grd.union.ll <- spTransform(grd.union, ll)
```

```{r, comment=NA}
llGRD <- GE_SpatialGrid(grd.union.ll)
llGRD_in <- over(llGRD$SG, grd.union.ll)
llSGDF <- SpatialGridDataFrame(
  grid = slot(llGRD$SG, "grid"), 
  proj4string = CRS(proj4string(llGRD$SG)), 
  data = data.frame(in0 = llGRD_in))
llSPix <- as(llSGDF, "SpatialPixelsDataFrame")

meuse_ll <- spTransform(meuse, CRS("+proj=longlat +datum=WGS84"))
llSPix$pred <- idw(log(zinc) ~ 1, meuse_ll, llSPix)$var1.pred
```

**Figure 4.3**. Interpolated log zinc ppm for the Meuse Bank data set shown in Google Earth

```{r, comment=NA}
png(file = "zinc_IDW.png", width = llGRD$width, height = llGRD$height, bg = "transparent")
par(mar = c(0, 0, 0, 0), xaxs = "i", yaxs = "i")
image(llSPix, "pred", col = bpy.colors(20))
dev.off()
kmlOverlay(llGRD, "zinc_IDW.kml", "zinc_IDW.png")
```

### 4.3.3 Other Import/Export Functions

## 4.4 Grass

```{r, comment=NA}
#system("g.version", intern = TRUE)
#gmeta6()
```

```{r, comment=NA}
#spear <- readRAST6(c("elevation.dem", "geology"), cat = c(FALSE, TRUE))
#summary(spear)
```

```{r, comment=NA}
#table(spear$geology)
#system("r.stats --q -cl geology", intern = TRUE)
```

**Figure 4.4**. Empirical cumulative distribution function of elevation for the Spearfish location

**Figure 4.5**. Boxplots of elevation by geology category, Spearfish location

```{r, comment=NA}
#bugsDF <- readVECT6("bugsites")
#vInfo("streams")
#streams <- readVECT6("streams", type = "line,boundary", remove.duplicates = FALSE)
```

```{r, comment=NA}
#summary(bugsDF)
```

### 4.4.1 Broad Street Cholera Data

```{r, comment=NA}
sohoSG <- readGDAL("sohoSG.tif")
names(sohoSG) <- c("snowcost_broad", "snowcost_not_broad")
buildings <- readOGR(".", "buildings", integer64 = "allow.loss")
proj4string(sohoSG) <- CRS(proj4string(buildings))
```

```{r, comment=NA}
deaths <- readOGR(".", "deaths", integer64="allow.loss")
names(deaths) <- c("cat", "long", "lat", "Num_Cases", "snowcost_broad", 
                   "snowcost_not_broad", "b_nearer")
o <- over(deaths, sohoSG)
deaths <- spCbind(deaths, o)
deaths$b_nearer <- deaths$snowcost_broad < deaths$snowcost_not_broad

by(deaths$Num_Cases, deaths$b_nearer, sum)
```


```{r, comment=NA}
oopar <- par(mfrow = c(1, 2), mar = c(5, 3, 1, 1) + 0.1)
b_wid <- table(deaths$b_nearer)
boxplot(snowcost_broad ~ b_nearer, deaths, width = b_wid, ylim = c(0, 450),
        ylab = "distance", xlab = "Broad Street", col = grey.colors(1, 0.8, 0.8, 2.2))
boxplot(snowcost_not_broad ~ b_nearer, deaths, width = b_wid, ylim = c(0, 450), 
        xlab = "Other pump", col = grey.colors(1, 0.8, 0.8, 2.2))
par(oopar)
```

**Figure 4.6**. Comparison of walking distances from homes of fatalities to the Broad Street pump or another pump by whether the Broad Street pump was closer or not

```{r, comment=NA}
nb_pump <- readOGR(".", "nb_pump")
b_pump <- readOGR(".", "b_pump")
```

```{r, comment=NA}
oopar <- par(mar = c(1, 1, 1, 1) + 0.1)
image(sohoSG, "snowcost_broad", breaks = seq(0, 750, 50), 
      col = colorRampPalette(brewer.pal(7, "Reds"))(15))
plot(buildings, col = "white", add = TRUE)
plot(buildings, angle = 45, density = 10, col = "grey70", add = TRUE)
symbols(coordinates(deaths), circles = 4*sqrt(deaths$Num_Cases), inches = FALSE, 
        add = TRUE, bg = c("brown2", "grey40")[deaths$b_nearer + 1])
rect(528900, 180550, 529040, 180990, border = NA, col = "white")
text(528970, 180950, "metres from\nBroad Street\npump", cex = 0.6)
legend.krige(c(528930, 528960), c(180600, 180900),
             sohoSG$snowcost_broad, vertical = TRUE, breaks = seq(0, 750, 50),
             col = colorRampPalette(brewer.pal(7, "Reds"))(15))
plot(nb_pump, add = TRUE, pch = 8, cex = 1.3, lwd = 2)
plot(b_pump, add = TRUE, pch = 4, cex = 1.5, lwd = 8, col = "white")
plot(b_pump, add = TRUE, pch = 4, cex = 1.5, lwd = 6)
rect(528900, 181330, 529140, 181380, border = NA, col = "white")
legend(c(528910, 529100), c(181350, 181380),
       legend = c("Broad Street pump", "other pumps"), pch = c(4, 8), bty = "n", 
       cex = 0.6, y.inter = 0.7)
rect(528900, 181270, 529180, 181335, border = NA, col = "white")
legend(c(528910, 529100), c(181275, 181325), 
       legend = c("nearer Broad Street pump", "nearer other pump"), 
       fill = c("grey40", "brown2"), bty = "n", cex = 0.6, y.inter = 0.7)
box()
par(oopar)
```

**Figure 4.7**. The 1854 London cholera outbreak near Golden Square

## 4.5 Other Import/Export Interfaces

### 4.5.1 Analysis and Visualization Applications

```{r, comment=NA}
#sp2Mondrian(scot_BNG, "scot_BNG.txt")
```

**Figure 4.8**. Screen shot of two linked Mondrian plots: a map of the Empirical Bayes smooth and a parallel boxplot for four variables, with the selected districts on the map (three northern mainland counties, Outer Hebrides, Orkney, and Shetland) split out as overlay boxplots

### 4.5.2 TerraLib and aRT

### 4.5.3 Other GIS and Web Mapping Systems

## 4.6 Installing rgdal
